---
formats: 
pdf:
title: "Differential Water Loss from Adaxial and Abaxial Petal Surfaces Suggests an Adaptive Role for Nyctinastic Movements in *Eschscholzia californica*"
author:  Kevin Simonin, Alyssa Miller, Annie Dang, Julia Rolls, Cameron Miller, Sean Britain, Kanchan Gautam, Griffin Wetzel, Lucy Gworek, Nicholas Rackard-Hilt and Gretchen LeBuhn

editor: visual
bibliography: references.bib
---

**Abstract**

\

**Introduction**

Plants employ diverse strategies to optimize reproductive success while minimizing the physiological cost of reproduction \[Ashman 1994, Schoen Asman 1995, Rody 2019, Galen \]. Although flowers represent a substantial resource investment, they are unique among plant organs in their relatively low dry matter content, relying largely on a hydrostatic skeleton rather than carbon-intensive structures for support [@niklas1992; Roddy et al. 2023]. This reliance on water for maintaining floral display, combined with the general absence of stomata, presents unique challenges for regulating floral water balance and minimizing the overall water cost of reproduction \[Galen 2005; Roddy et al 2016, Roddy et al. 2019\].

While the majority of water loss through leaves occurs through stomata, flowers primarily lose water through cuticular conductance (Roddy et al 2023; Teixido & Valladares 2014 ). This means that the maintenance of turgor pressure, essential for floral display and function, is strongly influenced by cuticular water loss, which unlike stomata is unregulated in response to short term changes in atmospheric demand (Roddy et al. BiorXiv). Despite the importance of understanding controls over cuticular water loss and overall flower water balance, rates of water loss from petal surfaces and the factors controlling the water cost of flowering remain poorly characterized.

One aspect of floral structure and function that may provide short-term regulation of cuticular water loss, is nyctinasty— opening and closing of flowers through the rythmic movement of petals that persist over multiple days. While the adaptive significance of these petal movements remains debated, mounting evidence suggests they may serve multiple functions, including pollinator attraction and protection from environmental stresses [@vandoorn2003]. We propose that nyctinastic movement of petals may additionally function as a water conservation mechanism during the critical period of reproductive display.

Nyctinastic movements show considerable variation in timing and duration across species, typically regulated by environmental cues such as light and temperature (**add reference**). These movements involve complex cellular mechanisms including ion transport and differential cell expansion, suggesting both energetic and water costs associated with repeated opening and closing (van Doorn and van Meeteren 2003).

*Eschscholzia californica* (California poppy) provides an ideal system for investigating the water conservation hypothesis of nyctinastic movements. This species exhibits pronounced daily flower opening in mid-morning and closing in mid-afternoon, with individual flowers persisting for three to five days. The large, bowl-shaped corolla presents extensive surface area to environmental factors that may influence water loss, including solar radiation and wind.

Here, we test the hypothesis that nyctinastic closure reduces water loss by comparing transpiration rates between adaxial and abaxial petal surfaces. We predict that the adaxial surface, which is protected during closure, will show higher rates of water loss than the abaxial surface, suggesting that flower closure serves to protect surfaces most vulnerable to water loss.\

**Methods**

Measruements were made on *Eschscholzia californica* (California poppy) individuals at several locations throughout the SF Bay area. *E. californica* is a widespread herbaceous plant native to the western United States, characterized by large, showy flowers that open late morning and close mid-afternoon. During closure, the petals fold inward forming a tight spiral that completely encloses the reproductive structures. This daily cycle of opening and closing continues for three to five days after which the petals abscise and fruit development begins. The species' adaptation to Mediterranean climates, where water availability can be unpredictable during the flowering season, makes it an ideal system for investigating water conservation strategies in flowers.

***Petal and Ovary Temperatures***

*In situ* measurements of petal and ovary temperatures were made on one flower from each of two *Eschscholzia californica* plants in Corte Madera (Marin county), California (38.0197° N 122.5022° W). On May 22-24,2022, the temperature inside the flower was measured using a Perfect Prime TL0201, 0.1mm diameter, K-Type Sensor Probe for K-Type thermocouple.  Single probes were placed touching the surface of the petals. Plant 1 was a control and plant 2 was held open by inserting a metal ring into the corolla.

On May 29-31, 2022, the temperature inside and outside the flower was measured using a Perfect Prime TL0201, 0.1mm diameter, K-Type Sensor Probe for K-Type thermocouple.  Single probes were placed touching the surface of the petal as well as being inserted into the ovary.  A fifth probe was placed adjacent to the flower to measure air temperature.  Probes were left in for 48 hours and data was recorded on a CR-6. Temperature was collected every 5 minutes. Thermocouples 1 and 4 were inserted in the ovary. Thermocouples 2 and 5 were suspended in the bowl. Thermocouple 3 was in open air. Data from the second flower on the second day was not collected as the petals abscised after the first 24 hours.

***Daytime Petal Water Content***

To examine the effect of time of day and floral opening on water content of flowers, we sampled one *Eschscholzia californica* petal from each of 30 different flowers per date on June 23, 26th and 28th , July 6th and August 10, 2022 located on San Francisco State University campus (add lat and long). No more than two flowers were sampled per individual plant.  Petals were held in a plastic container to minimize water loss until they could be weighted and photographed.  In the laboratory, each petal was photographed on a blue background next to a post-it used to calibrate ImageJ. A clear plastic sheet the size of standard lined paper (21.6 x 27.9 cm) was placed over the petal to flatten the petal to standardize measurements of area. Each reference photo was uploaded to ImageJ to estimate petal area.

Petal samples were weighed individually on a Mettler Toledo AX204 balance. After individual weight was taken (“Wet weight”), each petal was placed into a small coin envelope and dried in an Fisher Scientific Isotemp Standard Oven 600 series at 60℃ for at least 48 hours. After drying each individual, petal was weighed on the same balance to obtain the dry weight of each petal.  Total petal water content was estimated as Wet weight - Dry weight.  Petal water content per gram was estimated as Total water content/Dry weight. Petal water content per area was estimated as Total petal water content/Petal area.

***Petal Surface Conductance***

In situ measurements of petal surface conductance were made on *Eschscholzia californica* plants from randomly selected locations within the greater San Francisco Bay area. Petal surface conductance measurements were taken from 54 individuals using a Li-Cor Li-600 Porometer/Fluorometer. For each plant sampled, one petal per poppy flower was randomly selected for six measurements. Three measurements were taken from the abaxial surface and three measurements were taken from the adaxial surface. Surface conductance measurements were taken during anthesis when flowers were fully opened.

***Petal Surface Area***

All *Eschscholzia californica* flowers used for surface area measurements were sampled on the San Francisco State University campus grounds. 16 closed E. californica flowers were selected and marked to collect surface area measurements, as well as to return to when open. Closed flower measurements were taken between 7am-8:30am with the surface area of closed flowers calculated as:

$$A = \pi r(r + \sqrt{h^2 + r^2}) - \pi r^2$$

where *r* is the radius measured at the base of the petals and *h* is the height measured from the base to tip of the petals. \[add picture for reference\].

Marked *E. californica* flowers were revisited when open to collect and measure open surface area. Open surface area measurements were made digitally. For each flower, each individual petal was removed from its base and scanned for projected surface area using a document scanner (Cannon....). A sheet of glass was placed on top of the flower petals, flattening them to ensure the entire surface of the petals would be visible when scanning. Images were analyzed for surface area using ImageJ. Images were converted into binary and analyzed for surface area (cm^2^).

***Lifetime Petal Water Use***

We created a lifetime petal water use model to represent the variances of water use of E. californica flowers depending on the duration of floral opening and closing throughout their lifespan: SA ᐧ gsw ᐧ VPD ᐧ Time. SA represents surface area (m^2^), calculated above. G~sw~ represents mean surface conductance (mol m^-2^ s^-1^), obtained from the data collected from the Li-Cor Li-600 Porometer/Fluorometer. Average gsw was taken for both the abaxial and adaxial side of the flower petal. VPD, or vapor pressure deficit (kPa), was made as a constant of 1, removing the possibility of variability caused by differing VPD. Our time (s) variable used metadata collected from \[insert source\], finding that the average lifespan of E. californica flowers was 4 days. Within a 24 hour period, the flowers were closed for roughly 19 hours and open for 5 hours.

***Statistical Analysis***

All statistical analyses were done using RStudio.

**Results**

***Ovary and flower temperature***

Ovary and petal surface temperatures when the flower was closed were not significantly lower than ambient temperature.

NB:Data were collected from two flowers for 24 hours. Thermocouples 1 and 4 were inserted in the ovary. Thermocouples 2 and 5 were suspended in the bowl. Thermocouple 3 was in open air.

```{r}
#| label: Petal_temp_analysis
#| echo: false
#| warning: false
#| 
# Import required R packages.

#library("googlesheets4")
library(reshape2)
library(dplyr)
library(broom)
library(ggplot2)
library(nlme)
library(lubridate) # for working with dates
library(scales)   # to access breaks/formatting functions
library(gridExtra) # for arranging plots
library(tidyr) # for reshaping data
library(stringr)



#read in data and classify date variable
ov_temp = read.csv("~/GitHub/CApoppy/data/raw/fl_temp/cr6_22_may_22.csv", header = T, stringsAsFactors=FALSE)
ov_temp2 = read.csv("~/GitHub/CApoppy/data/raw/fl_temp/cr6_Temp.31.May.22.csv", header = T, stringsAsFactors=FALSE)

ov_temp$DateTime<- as.POSIXct(ov_temp$TIMESTAMP, format="%m/%d/%y %H:%M", tz = "UTC")
ov_temp2$DateTime<- as.POSIXct(ov_temp2$TIMESTAMP, format="%m/%d/%y %H:%M", tz = "UTC")


#Subset the data set to isolate the second day for the first flower
ov_temp3 <- subset(ov_temp2, RECORD > 328 & RECORD < 624) 
ov_temp4 <-subset(ov_temp2, RECORD > 391 )


df3 <- ov_temp3 %>%
  select(DateTime, Temp_C_Avg.1., Temp_C_Avg.2., Temp_C_Avg.3., Temp_C_Avg.4., Temp_C_Avg.5.) %>%
  gather(key = "sensor", value = "value", -DateTime)



AllTemp3 <- ggplot(df3, aes(DateTime, value)) +
           geom_line(aes(colour=sensor, group=sensor)) + # colour, group both depend on variable
           geom_point(aes(colour=sensor),                # colour depends on variable
               size=1)+
         ggtitle("Figure 1. Flower and ovary temperatures on two flowers, ovary and bowl") +
         xlab("Date") + ylab("Temperature (C)")

#render plot
AllTemp3


```

The second graph isolates the first flower where data was recorded continuously for 48 hours. Thermocouple 1 is in the ovary, Thermocouple 2 is touching the surface of an inner flower petal. Thermocouple 3 was in open air.

There was no significant difference in the ovary temperatures at night between the petal surface, ovary and air temperature.

```{r}
#| echo: false
#| warning: false

# Second day of blooming 
df4 <- ov_temp4 %>%
  select(DateTime,  Temp_C_Avg.1., Temp_C_Avg.2., Temp_C_Avg.3. ) %>%
  gather(key = "sensor", value = "value", -DateTime)


AllTemp4 <- ggplot(df4, aes(DateTime, value)) +
           geom_line(aes(colour=sensor, group=sensor)) + # colour, group both depend on variable
           geom_point(aes(colour=sensor),                # colour depends on variable
               size=1)+
         ggtitle("Figure 2. Flower and ovary temperature - one flower ovary and bowl") +
         xlab("Date") + ylab("Temperature (C)")

#render plot

AllTemp4
```

These data track temperatures from themocouples inserted into a flower that was closed and one that was open. This represents two nights on the first night, we held on flower open using wire.

```{r}
#| echo: false
#| warning: false

df1 <- ov_temp%>%
  select(DateTime, Temp_C_Avg.3., Temp_C_Avg.4., Temp_C_Avg.5. ) %>%
  gather(key = "variable", value = "value", -DateTime)
head(df1, 3)

AllTemp1 <- ggplot(df1, aes(DateTime, value)) +
           geom_line(aes(colour=variable, group=variable)) + # colour, group both depend on variable
           geom_point(aes(colour=variable),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min May 27-28, 2022") +
         xlab("Date") + ylab("Air Temperature (C)")



df2 <- ov_temp %>%
  select(DateTime,  Temp_C_Avg.3., Temp_C_Avg.1., Temp_C_Avg.2. ) %>%
  gather(key = "variable", value = "value", -DateTime)
#head(df1, 3)

AllTemp2 <- ggplot(df2, aes(DateTime, value)) +
           geom_line(aes(colour=variable, group=variable)) + # colour, group both depend on variable
           geom_point(aes(colour=variable),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min May 27-28, 2022") +
         xlab("Date") + ylab("Air Temperature (C)")



#render plot
AllTemp1
AllTemp2
```

***Daytime Petal Water Content***

```{r}
#| label: Petal_water_analysis
#| warning: false
#| include: false


# Define paths to data sets. If you don't keep your data in the same directory as the code, adapt the path names.

data <- read.csv("~/GitHub/CApoppy/data/raw/Petal_water.csv", header=TRUE, stringsAsFactors=TRUE)

data<-data[-c(1,2,3,4),]

data$Time_cat <- as.factor(data$Time_cat)
#remove outlier
data <- data[-c(76),]
data <- data[-c(79),]
```

Total water content of petals is highest before opening in the morning (06:00), declines from the time of opening (9:00) to time of closing (15:00). Upon closing (15:00) water content stabilizes and begins to increase.

```{r}
#| echo: false
#| warning: false
plot1<- ggplot(data, aes(group = Time_cat, y = water_mg, fill = Time_cat)) + 
  stat_boxplot(geom = "errorbar",
               width = 0.75) + 
  geom_boxplot() +
        ggtitle("Figure 3. Total water content (mg) of petals at different times of day") +
        xlab("Time collected") + ylab("Water (mg)") +
        theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

plot1
```

Water content per unit surface area is highest before the flower opens (06:00), drops as soon as the flower has opened (09:00) and remains lower while the petals are open and then increases when the flower closes (15:00)

```{r}
#| echo: false
#| warning: false
plot2<- ggplot(data, aes(group = Time_cat, y = water_cm2, fill = Time_cat)) + 
  stat_boxplot(geom = "errorbar",
               width = 0.75) + 
  geom_boxplot() +
         ggtitle("Figure 4. Petal water content per unit area (cm2) of petals at different times of day") +
         xlab("Time collected") + ylab("Water (cm2)")+
        theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

plot2
```

Upon opening (06:00) and when petals close (15:00) , petal water content per unit area is higher.

***Lifetime petal water use***

The contribution of petals to total flower water use was significantly influenced by differences in abaxial and adaxial petal surface conductance and repeated flower opening and closing. The average surface conductance (*g*~sw~) to water vapor for the adaxial petal surface was \~ 5 times greater than the abaxial surface (Fig ?, adaxial *g*~sw~ = 0.011 moles m^2^ s^-1^, abaxial *g*~sw~ = 0.0024 moles moles m^2^ s^-1^, n = 54 p \< 0.05).

```{r}
## read in LI 600 data
DataA <- read.csv("Petals_KevinSimonin_2024_03_17.csv")
DataB <- read.csv("Petals_AnnieDang_2024_03_08.csv")
DataC <- read.csv("Petals_KanchanGautam_2024_03_18.csv")
DataD <- read.csv("Petals_CamMiller_2024_03_15.csv")
DataE <- read.csv("Petals_JR_2024_03_08.csv")
DataF <- read.csv("Lucy_gsw_2022.csv")
DataG <- read.csv("Petal_water.csv")
DataH <- read.csv("Ovary_temps_June2.csv")

##merge LI 600 data files
Data <- DataA %>%
  full_join(DataB) %>%
  full_join(DataC) %>%
  full_join(DataD) %>%
  full_join(DataE) %>%
  full_join(DataF)


## keep relevant columns and place in a new Dataframe, grouped by Plant
Data_1 <- Data %>% 
  group_by(Plant) %>%
  select(Time, Date, Species, t_or_b, Plant, Location, gsw)


## Create a data frame for petal bottom
Data_1_b <- filter(Data_1, t_or_b == "b")


##Create a dataframe of mean gsw for petal bottom
Data_gs_bottom <- Data_1_b %>%
  summarise(gsw = mean(gsw))

Data_gs_bottom$t_or_b <- "Bottom"

Data_gs_bottom

## Create a data frame for petal top
Data_1_t <- filter(Data_1, t_or_b == "t")

##Create a dataframe of mean gsw for petal top
Data_gs_top <- Data_1_t %>%
  summarise(gsw = mean(gsw))

Data_gs_top$t_or_b <- "Top"

Data_gs_top

##merge data frames with mean gsw for top and bottom

Data_gs_t_and_b <- full_join(Data_gs_bottom, Data_gs_top)


Data_gs_t_and_b


##summarize data
Data_gs_t_and_b %>%
  group_by(t_or_b) %>%
  get_summary_stats(gsw, type = "mean_sd")

by(Data_gs_t_and_b,
   INDICES = Data_gs_t_and_b[, "t_or_b"],
   FUN = summary)

##Paired t-test
Ttest <- Data_gs_t_and_b %>%
  t_test(gsw ~ t_or_b, paired = TRUE) %>%
  add_significance()
Ttest

##Figure

AB_AD_gsw_boxplot <- ggplot(Data_gs_t_and_b, aes(factor(t_or_b), gsw, fill=t_or_b)) + 
  geom_boxplot(color = "black", fill = "white") +
  ylab(expression(paste("Surface conductance (mmol m"^-2, " s"^-1,")"))) +
  theme_classic2() +
  theme(axis.title.x = element_blank())
  
AB_AD_gsw_boxplot
```

We incorporated the difference in abaxial and adaxial *g*~sw~ into a life time flower water use model to evaluate the effects of variable *g*~sw~ and nyctinasty on life time flower water use (Fig ?).

```{r}
## read in petal surface area data
Petal_surface_area <- read.csv("Petal_surface_area.csv")


## calculate petal surface area when flowers are closed
Petal_surface_area <- Petal_surface_area %>%
  mutate(diameter_cm = diameter_inch*2.54) %>%
  mutate(length_cm = length_inch*2.54) %>%
  mutate(radius_cm = diameter_cm/2) %>%
  mutate(closed_area_cm2 = pi*radius_cm*(radius_cm+sqrt(length_cm^2 + radius_cm^2)) - pi*radius_cm^2)

#removing data with missing datapoints
Petal_surface_area <- subset(Petal_surface_area, !is.na(open_area_cm2))

## model water loss using mean gsw for top and bottom surface
## 0.0001 = conversion from cm2 to m2
## 86400 = number of seconds in a day
## 4 = number of days

Petal_surface_area <- Petal_surface_area %>%
  mutate(gsw_top = 0.011464) %>%
  mutate(gsw_bottom = 0.0024149) %>%
  mutate(E_closed_gsw_bottom_mol = gsw_bottom * closed_area_cm2 * 0.0001 * 86400 *4) %>%
  mutate(E_open_gsw_bottom_top_mol = (gsw_bottom + gsw_top) * open_area_cm2 * 0.0001 * 86400 * 4) %>%
  mutate(E_open_closed_mol = (68400 * closed_area_cm2 * 0.0001 * gsw_bottom)+
           (18000 * open_area_cm2 * 0.0001 * (gsw_bottom + gsw_top))*4) %>%
  mutate(E_closed_gsw_top_mol = gsw_top * closed_area_cm2 * 0.0001 * 86400 * 4) %>%
  mutate(E_open_gsw_top_only_mol = gsw_top* 2 * open_area_cm2 * 0.0001 * 86400 * 4) %>%
  mutate(E_open_gsw_bottom_only_mol = gsw_bottom * 2 * open_area_cm2 * 0.0001 * 86400 * 4) %>%
  mutate(E_open_closed_gsw_top_only_mol = ((gsw_top * closed_area_cm2 * 0.0001 * 68400) +
                                             (gsw_top * 2* open_area_cm2 * 0.0001 * 18000))*4) %>%
  mutate(E_open_closed_gsw_bottom_only_mol = ((gsw_bottom * closed_area_cm2 * 0.0001 * 68400) +
                                             (gsw_bottom * 2* open_area_cm2 * 0.0001 * 18000))*4)

#organize data such that model is a descriptor, this will make graphing and data analysis much easier


Petal_SA_1 <- Petal_surface_area %>%
  group_by(plant) %>%
  select(plant, E_closed_gsw_bottom_mol, E_closed_gsw_top_mol, E_open_gsw_bottom_only_mol,
         E_open_gsw_top_only_mol, E_open_gsw_bottom_top_mol, E_open_closed_gsw_bottom_only_mol,
         E_open_closed_gsw_top_only_mol, E_open_closed_mol)

Petal_SA_1_closed_gsw_bottom <- Petal_SA_1 %>%
  group_by(plant) %>%
  select(plant, E_closed_gsw_bottom_mol)

Petal_SA_1_closed_gsw_bottom <- Petal_SA_1_closed_gsw_bottom %>%
  mutate(E_mol = E_closed_gsw_bottom_mol) %>%
  mutate(model = "closed_gs_bottom")

Petal_model1 <- Petal_SA_1_closed_gsw_bottom %>%
  select(plant,model, E_mol)

Petal_SA_1_closed_gsw_top <- Petal_SA_1 %>%
  select(plant, E_closed_gsw_top_mol)

Petal_SA_1_closed_gsw_top <- Petal_SA_1_closed_gsw_top %>%
  mutate(E_mol = E_closed_gsw_top_mol) %>%
  mutate(model = "closed_gs_top")

Petal_model2 <- Petal_SA_1_closed_gsw_top %>%
  select(plant,model, E_mol)

Petal_SA_1_open_gsw_bottom <- Petal_SA_1 %>%
  select(plant, E_open_gsw_bottom_only_mol)

Petal_SA_1_open_gsw_bottom <- Petal_SA_1_open_gsw_bottom %>%
  mutate(E_mol = E_open_gsw_bottom_only_mol) %>%
  mutate(model = "open_gs_bottom")

Petal_model3 <- Petal_SA_1_open_gsw_bottom %>%
  select(plant, model, E_mol)

Petal_SA_1_open_gsw_top <- Petal_SA_1 %>%
  select(plant, E_open_gsw_top_only_mol)

Petal_SA_1_open_gsw_top <- Petal_SA_1_open_gsw_top %>%
  mutate(E_mol = E_open_gsw_top_only_mol) %>%
  mutate(model = "open_gs_top")

Petal_model4 <- Petal_SA_1_open_gsw_top %>%
  select(plant, model, E_mol)

Petal_SA_1_open_gsw_bottom_top <- Petal_SA_1 %>%
  select(plant, E_open_gsw_bottom_top_mol)

Petal_SA_1_open_gsw_bottom_top <- Petal_SA_1_open_gsw_bottom_top %>%
  mutate(E_mol = E_open_gsw_bottom_top_mol) %>%
  mutate(model = "open_gs_bottom_top")

Petal_model5 <- Petal_SA_1_open_gsw_bottom_top %>%
  select(plant, model, E_mol)

Petal_SA_1_open_closed_gsw_bottom_only_mol <- Petal_SA_1 %>%
  select(plant, E_open_closed_gsw_bottom_only_mol)

Petal_SA_1_open_closed_gsw_bottom_only_mol <- Petal_SA_1_open_closed_gsw_bottom_only_mol %>%
  mutate(E_mol = E_open_closed_gsw_bottom_only_mol) %>%
  mutate(model = "open_closed_gs_bottom")

Petal_model6 <- Petal_SA_1_open_closed_gsw_bottom_only_mol %>%
  select(plant, model, E_mol)

Petal_SA_1_open_closed_gsw_top_only_mol <- Petal_SA_1 %>%
  select(plant, E_open_closed_gsw_top_only_mol)

Petal_SA_1_open_closed_gsw_top_only_mol <- Petal_SA_1_open_closed_gsw_top_only_mol %>%
  mutate(E_mol = E_open_closed_gsw_top_only_mol) %>%
  mutate(model = "open_closed_gs_top")

Petal_model7 <- Petal_SA_1_open_closed_gsw_top_only_mol %>%
  select(plant, model, E_mol)

Petal_SA_1_open_closed_actual <- Petal_SA_1 %>%
  select(plant, E_open_closed_mol)

Petal_SA_1_open_closed_actual <- Petal_SA_1_open_closed_actual %>%
  mutate(E_mol = E_open_closed_mol) %>%
  mutate(model = "open_closed_actual")

Petal_model8 <- Petal_SA_1_open_closed_actual %>%
  select(plant, model, E_mol)

##merge individual model outputs into a single file

Model_E_Data <- Petal_model1 %>%
  full_join(Petal_model2) %>%
  full_join(Petal_model3) %>%
  full_join(Petal_model4) %>%
  full_join(Petal_model5) %>%
  full_join(Petal_model6) %>%
  full_join(Petal_model7) %>%
  full_join(Petal_model8)

##summarize data
Model_E_Data %>%
  group_by(model) %>%
  get_summary_stats(E_mol, type = "mean_sd")

one.way_E_mol <- aov(E_mol ~ model, data = Model_E_Data)
summary(one.way_E_mol)

tukey.test.E_mol <- TukeyHSD(one.way_E_mol)
tukey.test.E_mol

boxplot <- ggplot(Model_E_Data, aes(factor(model), E_mol, fill=model)) + 
  geom_boxplot() +
  scale_fill_brewer(palette="Dark2") +
  ylab(expression(paste("Total E (mol)"))) +
  theme_classic2()
theme(axis.title.x = element_blank())

boxplot


```

Differences in surface conductance between the abaxial and adaxial petal surfaces along together with floral opening and closing significantly influenced lifetime flower water use (F = , p \< 0.05 \##**Need to talk about the statistical analysis on this.**)

\*\* References\*\*

### **Discussion**

Nyctinastic petals have been shown / hypotehsized to benefit the liftime reproductive effort of plants in a number of ways. Here we show strong evidence for the adaptive benefits of nyctinastic petal movements with respect to the lifetime water use of *Eschscholzia californica* flowers and by extension whole plant water use. Further, the data presented here highlights the importance of characterizing the surface conductance of both abaxial and adaxial petal surfaces when evaluating lifetime flower water use of nyctinastic flowers.

Previous work ##Here provide some recaps of shown / hypothesized benefits that relate to temperature, and then bring in our results.

##Move toward discussing the petal water content data and lifetime flower water use.

Over the course of a day we observed significant (need to check stats) diurnal variaiton in the water balance of flowers. During periods when flowers were open there was a decrease in petal water storage, e.g. petal water content per unit surface area. The observed decrease in water storage when flowers were open suggests a transient increase in water loss relative to water supply as compared to periods when flowers are closed. There are two major factors that contribute to total flower water loss that are altered by nyctinastic petals. Part of the transient disequilibrium between water loss and supply can be attributed to the amount of petal surface area directly exposed to the broader atmosphere when flowers are open relative to being closed (Fig. ?, open versus closed surface area). A closed flower leads to a decoupling between the atmosphere directly above the adaxial surface and the broader atmosphere resulting in a reduce rate of water loss; functionaly a lower adaxial boundary layer conductance given that flowers are not fully sealed when closed. Another factor, is the potential for variation in the surface conductance of adazial and abaxial surfaces. For Eschscholzia californica petals the adaxial petal surface showed a significantly higer surface conductance to water vapor than the abaxial surface.

Modelling results suggest that the lifetime water use of nyctinastic lowers is half the amount of water use of flowers that were to remain open all the time and equal to the total wter use of flowers that are open all the time if the adaxial surface conductance was lowered to that of the abaxial surface. This begs the question why does the adaxial surface of Eschalzia petals have such a high capacity to loose water as compared to the abaxial surface. Flowers could remain day and night and loose the same amount of water as nyctinastic flowers if the adaxial surface conductance was lower.

##Old text from the methods section

This nyctinastic behavior occurs even under constant light conditions, suggesting an endogenous rhythm that is entrained by environmental cues. The flowers lack a rigid support structure, instead relying on hydrostatic pressure to maintain their characteristic cup shape when open.
