---
title: "Cal Poppy ovary temp"
author: "Gretchen LeBuhn"
date: "5/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/gretchen/Library/CloudStorage/Box-Box/1. Lab research/CApoppy/data/fl_temp')
```

## R Markdown

These are data collected to track flowering time in Eschscholzia californica.  This represents two nights on the first night, we held on flower open using wire.

```{r set libraries and dir, echo=FALSE}
library(lubridate) # for working with dates
library(ggplot2)  # for creating graphs
library(scales)   # to access breaks/formatting functions
library(gridExtra) # for arranging plots
library(tidyr) # for reshaping data
library(dplyr)
library(stringr)


```

```{r, set dir, echo = FALSE}
#setwd('~/Dropbox/CApoppy/data')
#setwd('/Users/gretchen/Library/CloudStorage/Box-Box/1. Lab research/CApoppy/data/fl_temp')
#datadir="~/Dropbox/CApoppy/data"
```
I've modified the CR6 data by dropping the last header line and switching lines 2 and 3 so that line 2 becomes the header for the data.  I had to do text to columns and then save it as .csv for it to import correctly.  Data set ov_temp measures themocouples inserted into a flower that was closed and one that was open.  Data set ov_temp3 is two flowers each with a thermocouple in the ovary and one in the bowl.  Dataset ov_temp 4 is one flower so read just Temp 1 and 2 for the flower data.  The other 3 thermocouples were in the open.  Jun 4 night it rained until jun 5 at 11.
```{r, import data, echo = FALSE}

ov_temp = read.csv("cr6_Temp.28.May.22.csv", header = T, stringsAsFactors=FALSE)
ov_temp2 = read.csv("cr6_Temp.31.May.22.csv", header = T, stringsAsFactors=FALSE)
ov_temp_current= read.csv("cr6_Temp.5.Jun.22.csv", header = T, stringsAsFactors=FALSE)
#format date and time
ov_temp$DateTime<- as.POSIXct(ov_temp$TIMESTAMP, format="%m/%d/%y %H:%M", tz = "UTC")
ov_temp2$DateTime<- as.POSIXct(ov_temp2$TIMESTAMP, format="%m/%d/%y %H:%M", tz = "UTC")
ov_temp_current$DateTime<- as.POSIXct(ov_temp_current$TIMESTAMP, format="%m/%d/%y %H:%M", tz = "UTC")





#Subset the second data set
ov_temp3 <- subset(ov_temp2, RECORD > 328 & RECORD < 624) 
ov_temp4 <-subset(ov_temp2, RECORD > 391 )

```

## Plotting temperature data for the first few days

Using this plotting tutorial
https://www.neonscience.org/resources/learning-hub/tutorials/dc-time-series-plot-ggplot-r

```{r airtemp, echo=FALSE}

df1 <- ov_temp%>%
  select(DateTime, Temp_C_Avg.3., Temp_C_Avg.4., Temp_C_Avg.5. ) %>%
  gather(key = "variable", value = "value", -DateTime)
head(df1, 3)

AllTemp1 <- ggplot(df1, aes(DateTime, value)) +
           geom_line(aes(colour=variable, group=variable)) + # colour, group both depend on variable
           geom_point(aes(colour=variable),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min May 27-28, 2-22") +
         xlab("Date") + ylab("Air Temperature (C)")



df2 <- ov_temp %>%
  select(DateTime,  Temp_C_Avg.3., Temp_C_Avg.1., Temp_C_Avg.2. ) %>%
  gather(key = "variable", value = "value", -DateTime)
head(df1, 3)

AllTemp2 <- ggplot(df2, aes(DateTime, value)) +
           geom_line(aes(colour=variable, group=variable)) + # colour, group both depend on variable
           geom_point(aes(colour=variable),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min May 27-28, 2-22") +
         xlab("Date") + ylab("Air Temperature (C)")



#render plot
AllTemp1
AllTemp2
```
Now, I am plotting data from two flowers for 24 hours.  Thermocouples 1 and 4 were inserted in the ovary.  Thermocouples 2 and 5 were suspended in the bowl.  Thermocouple 3 was in open air.
The second graph isolates the first flower which continued to bloom for a second day.  Thermocouple 1 is in the ovary, Thermocouple 2 is suspended in the flower.  


```{r}

df3 <- ov_temp3 %>%
  select(DateTime, Temp_C_Avg.1., Temp_C_Avg.2., Temp_C_Avg.3., Temp_C_Avg.4., Temp_C_Avg.5.) %>%
  gather(key = "sensor", value = "value", -DateTime)
head(df1, 3)


AllTemp3 <- ggplot(df3, aes(DateTime, value)) +
           geom_line(aes(colour=sensor, group=sensor)) + # colour, group both depend on variable
           geom_point(aes(colour=sensor),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min May 29-30, 2022 4 flowers, ovary and bowl") +
         xlab("Date") + ylab("Air Temperature (C)")

df4 <- ov_temp4 %>%
  select(DateTime,  Temp_C_Avg.1., Temp_C_Avg.2., Temp_C_Avg.3. ) %>%
  gather(key = "sensor", value = "value", -DateTime)
head(df1, 3)

AllTemp4 <- ggplot(df4, aes(DateTime, value)) +
           geom_line(aes(colour=sensor, group=sensor)) + # colour, group both depend on variable
           geom_point(aes(colour=sensor),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min May 30-31, 2022- one flower ovary and bowl") +
         xlab("Date") + ylab("Air Temperature (C)")

#render plot
AllTemp3
AllTemp4
```

Now I am trying to plot by time using all the data.


```{r}

dfc <- ov_temp_current %>%
  select(TIMESTAMP, Temp_C_Avg.1., Temp_C_Avg.2., Temp_C_Avg.3., Temp_C_Avg.4., Temp_C_Avg.5.) %>%
  gather(key = "sensor", value = "value", -TIMESTAMP)
head(dfc, 3)

AllTempc <- ggplot(dfc, aes(TIMESTAMP, value)) +
           geom_line(aes(colour=sensor, group=sensor)) + # colour, group both depend on variable
           geom_point(aes(colour=sensor),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min , 2022") +
         xlab("Date") + ylab("Air Temperature (C)") +
         scale_x_discrete(guide = guide_axis(check.overlap = TRUE)) +
        theme(axis.text.x=element_text(angle=60, hjust=1))
AllTempc


#gather data from sensors 1 and 2
dfc2<-gather(ov_temp_current, key="sensor", value="value", 7:8)
head(dfc2, 3)

AllTempc2 <- ggplot(dfc2, aes(TIMESTAMP, value )) +
           geom_line(aes(colour=sensor, group=sensor)) + # colour, group both depend on variable
           geom_point(aes(colour=sensor),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min , 2022") +
         xlab("Date") + ylab("Air Temperature (C)") +
         scale_x_discrete(guide = guide_axis(check.overlap = TRUE)) +
        theme(axis.text.x=element_text(angle=60, hjust=1))
AllTempc2


#gather data from columns sensors 4 and 5
dfc3<-gather(ov_temp_current, key="sensor", value="value", 10:11)
head(dfc2, 3)

AllTempc3 <- ggplot(dfc3, aes(TIMESTAMP, value )) +
           geom_line(aes(colour=sensor, group=sensor)) + # colour, group both depend on variable
           geom_point(aes(colour=sensor),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 5 min , 2022") +
         xlab("Date") + ylab("Air Temperature (C)") +
         scale_x_discrete(guide = guide_axis(check.overlap = TRUE)) +
        theme(axis.text.x=element_text(angle=60, hjust=1))
AllTempc3

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
