---
author: CApoppy Project
execute:
  echo: true
  message: false
  warning: false
format:
  html:
    toc: true
title: Dummy Analysis Example
---

``` {r}
#| label: setup
#| include: false

# Load standardized project setup
source(here::here("R", "project_setup.R"))
```

# Purpose

This dummy file demonstrates:

-   Reading raw data
-   Performing a simple summary
-   Saving a figure to `outputs/figures`
-   Saving a table to `outputs/tables`
-   Rendering report to `outputs/reports`

# 1. Simulate or Read Data

In real scripts, you would read data like this:

``` r
# df <- read_csv(file.path(paths$data_raw, "your_folder", "your_file.csv"))
```

Here we simulate data:

``` {r}
set.seed(123)

df <- tibble(
  population = rep(c("Cojo", "PtCon"), each = 50),
  petal_area = rnorm(100, mean = 5, sd = 1),
  gsw = rnorm(100, mean = 0.3, sd = 0.05)
)

head(df)
```

# 2. Simple Summary Table

``` {r}
summary_tbl <- df %>%
  group_by(population) %>%
  summarise(
    mean_area = mean(petal_area),
    mean_gsw = mean(gsw),
    n = n(),
    .groups = "drop"
  )

summary_tbl
```

Save the table:

``` {r}
save_table_csv(summary_tbl, "dummy_summary_table.csv")
```

# 3. Create and Save a Figure

``` {r}
p <- ggplot(df, aes(petal_area, gsw, color = population)) +
  geom_point() +
  labs(
    x = "Petal Area",
    y = "Stomatal Conductance (gsw)",
    title = "Dummy Petal Area vs GSW"
  )

p
```

Save the figure:

``` {r}
save_fig(p, "dummy_petal_area_vs_gsw.png")
```

# 4. Save a Model Object

``` {r}
model <- lm(gsw ~ petal_area + population, data = df)
summary(model)
```

Save model object:

``` {r}
save_rds(model, "dummy_linear_model.rds")
```

# 5. Rendering the Report

From the project root, render this file with:

``` bash
quarto render scripts/00_dummy_example.qmd --output-dir outputs/reports
```
