---
title: "Poppy dry weight"
author: "Gretchen LeBuhn"
date: '2022-07-01'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Import required packages.

library(dplyr)
library(tidyr)
library(ggplot2)
#library(vioplot)
library(corrplot)
library(gmodels)
library(matrixStats)
library(reshape)
```

## California Poppy Petal data

We started to analyze data collected on California poppies from SFSU campus. We collected petals at two different times of day (9 am and 2 pm), weighed them, digitized a photo to get area and dried them for 48 hours to get dry weights.

```{r import data}

setwd('/Users/gretchen/Library/CloudStorage/Box-Box/1. Lab research/CApoppy/data')
weight <- read.csv("CApoppy.weight.csv", header=TRUE, stringsAsFactors=TRUE)
weight$uv.index <-as.factor(weight$uv.index)
weight$Plant<-as.factor(weight$Plant)
meltWeight <- melt(weight)

```

We started by looking at box plots for the variables. water_cm2 is the (Wet.weight - Dry.weight)/petal.size. tissue_cm2 is the Dry weight/petal.size. water.dry.wgt is the (Wet.weight-Dry.weight)/Dry.weight

```{r pressure, echo=FALSE}
###  Boxplots

ggplot(meltWeight,aes(x = Petal.status,y = value)) + facet_wrap(~variable, scales = "free_y") + geom_boxplot() 
```

Here we test for differences between the water_cm2, tissue_cm2 and water/dry weight

```{r, anovas of water and dry weight}
wilcox.test(water_cm2 ~ Petal.status, data = weight)
wilcox.test(tissue_cm2 ~ Petal.status, data = weight)
wilcox.test(water.dry.wgt ~ Petal.status, data = weight)


```

There is no significant difference between water or dry weight of petals at 9 am (closed) and 2 pm (open).
