---
title: "Comparison of gsw by Population and Leaf Surface"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(ggpubr)
```

## Load Data

This merges the Cojo and CojoHQ population data and adds 16 to the plant ids in the Cojo group so that the plant ids are different. The Cojo data were taken on April 19, 2025 and the CojoHQ data were taken on April 17, 2025. This code also filters out gsw values that are greater than 0.05 or less than -0.05.

```{r}
# Load the data
#| echo: false
#| #| warning: false

data <- read.csv("~/GitHub/CApoppy/data/processed/porometer_4pops.csv", skip = 1, header = TRUE)

# Clean the data: remove NA values and ensure factor types
data_clean <- data %>%
  mutate(Date = as.Date(Date)) %>%  # Convert Date column to Date type
  filter(!is.na(gsw), !is.na(Population), !is.na(Plant), !is.na(`Top.or.Bottom`)) %>%
  filter(gsw < 0.05, gsw > -0.05) %>%
  mutate(
    Population = case_when(
      Population %in% c("Cojo", "CojoHQ") ~ "Cojo",
      TRUE ~ as.character(Population)
    ),
    Plant = case_when(
  Date %in% as.Date(c("2025-04-19", "2025-04-20")) ~ Plant + 16,
  TRUE ~ Plant
),
    Population = factor(Population),
    TopBottom = factor(`Top.or.Bottom`)
  )

head(data_clean)
```

## Summary Statistics

```{r}
data_clean %>%
  group_by(Population, TopBottom) %>%
  summarise(
    mean_gsw = mean(gsw),
    sd_gsw = sd(gsw),
    n = n(),
    .groups = 'drop'
  )
```

## Histogram of gsw by Population and Leaf Surface

```{r}

# Overall histogram (all populations combined)
hist_overall <- ggplot(data_clean, aes(x = gsw)) +
  geom_histogram( color = "black", fill = "darkorange", bins=50) +
  geom_vline(aes(xintercept = mean(gsw, na.rm = TRUE)),
             color = "blue", linetype = "dashed", linewidth = 0.8) +
  labs(title = "Histogram of gsw (All Populations)", x = "gsw", y = "Count") +
  theme_minimal()

# Calculate means per TopBottom x Population group
means_df <- data_clean %>%
  group_by(TopBottom, Population) %>%
  summarize(mean_gsw = mean(gsw, na.rm = TRUE))

# Faceted histogram with means
hist_by_population <- ggplot(data_clean, aes(x = gsw)) +
  geom_histogram(color = "black", fill = "darkorange", bins = 30) +
  geom_vline(data = means_df, aes(xintercept = mean_gsw),
             color = "blue", linetype = "solid", size = 0.6) +
  labs(title = "Histogram of gsw by Population and Leaf Surface",
       x = "gsw", y = "Count") +
  facet_grid(TopBottom ~ Population) +
  theme_minimal()

# Show the overall histogram
print(hist_overall)

# Show histograms for each population
print(hist_by_population)
```

## Boxplot of gsw by Population and Leaf Surface

```{r}
ggplot(data_clean, aes(x = Population, y = gsw, fill = TopBottom)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(
    title = "gsw by Population and Leaf Surface",
    x = "Population", y = "gsw", fill = "Leaf Surface"
  ) +
  scale_fill_manual(values = c("Top" = "springgreen", "Bottom" = "darkorange" )) +
  theme_minimal()
```

## Two-Way ANOVA

```{r}
anova2 <- aov(gsw ~ Population * TopBottom, data = data_clean)
summary(anova2)
```

## Post-Hoc Comparison (Tukey HSD)

```{r}
# Tukey post-hoc for interaction effects
TukeyHSD(anova2)
```

## Interaction Plot

```{r}
gginteraction <- ggplot(data_clean, aes(x = TopBottom, y = gsw, color = Population, group = Population)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(title = "Interaction Plot: Population × Leaf Surface", x = "Leaf Surface", y = "Mean gsw") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1) +
  labs(
    title = "Interaction Plot: Population × Leaf Surface (with SE)",
    x = "Leaf Surface",
    y = "Mean gsw ± SE"
  ) +
  theme_minimal()

gginteraction
```

Separate figures for each population

```{r}


gginteraction2 <- ggplot(data_clean, aes(x = TopBottom, y = gsw, group = Population)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1) +
  labs(
    title = "Interaction Plot: Population × Leaf Surface (with SE)",
    x = "Leaf Surface",
    y = "Mean gsw ± SE"
  ) +
  facet_wrap(~ Population) +  # Creates one plot per population
  theme_minimal()

gginteraction2
```

```{r}
library(broom)

library(dplyr)
library(tidyr)
library(purrr)
library(broom)



t_test_results <- data_clean %>%
  filter(TopBottom %in% c("Top", "Bottom")) %>%
  select(Population, Plant, TopBottom, gsw) %>%
  pivot_wider(names_from = TopBottom, values_from = gsw) %>%
  filter(!is.na(Top) & !is.na(Bottom)) %>%
  group_by(Population) %>%
  summarise(
    n = n(),  # number of plants with both Top and Bottom values
    t_test = list(t.test(Top, Bottom, paired = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(tidy_result = map(t_test, tidy)) %>%
  unnest(tidy_result) %>%
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "n.s."
    )
  ) %>%
  select(Population, n, estimate, statistic, p.value, sig, conf.low, conf.high)

print(t_test_results)
```
