---
title: "plant Data Analysis"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(knitr)
library(broom)
```

## Data Import and Cleaning

First, we'll import the three datasets and clean them to prepare for merging.

```{r import-data}
# Import datasets
petal_data <- read_csv("../data/raw/plant/petal_weight.csv", na = c("", "NA", "NULL")) %>%
  filter(!is.na(population)) # Remove rows with missing population data

plant_traits <- read_csv("../data/raw/plant/plant_traits.csv", na = c("", "NA", "NULL"))

# For porometer data, skip the first row which contains headers
porometer <- read_csv("../data/processed/porometer_4pops.csv", skip = 1, na = c("", "NA", "NULL")) %>%
  # Drop data from April 19 and April 20
  filter(date != "2025-04-19" & date != "2025-04-20"& date != "2025-04-21")
```

Let's inspect the data structure of each dataset:

```{r inspect-data}
# Inspect first few rows
glimpse(petal_data)
glimpse(plant_traits)
glimpse(select(porometer, "Obs#", time, date, population, plant, gsw)) # Show relevant columns

# Check for unique populations in each dataset
cat("Unique populations in petal_data:", paste(unique(petal_data$population), collapse = ", "), "\n")
cat("Unique populations in plant_traits:", paste(unique(plant_traits$population), collapse = ", "), "\n")
cat("Unique populations in porometer:", paste(unique(porometer$population), collapse = ", "), "\n")
```

## Data Preparation

We need to standardize the datasets before merging: 1. Fix date formats 2. Standardize population names 3. Rename columns for consistency

```{r prepare-data}
# Clean and standardize petal weight data
petal_clean <- petal_data %>%
  # Convert plant to numeric if not already numeric
  mutate(plant = as.numeric(plant),
         # Ensure date is in proper format
        date = as_date(date)) %>%
  # Standardize population names
  mutate(
    # First fix the specific Pt. Conception issue
    population = ifelse(population == "Pt. Conception", "Pt_Conception", population),
    # Then handle other standardizations
    population = str_replace(population, "\\.", "_"),
    # Finally handle specific populations
    population = case_when(
      population == "Cojo" | population == "CojoHQ" ~ "Cojo",
      TRUE ~ population
    ))

# Clean and standardize plant traits data
plant_traits_clean <- plant_traits %>%
  # Convert date format - handle DD-MMM-YY format
  mutate(date = dmy(date),
         # Rename individual to plant for consistency
         plant = individual) %>%
  # Standardize population names
  mutate(
    # First fix the specific Pt. Conception issue
    population = ifelse(population == "Pt. Conception", "Pt_Conception", population),
    # Then handle other standardizations
    population = str_replace(population, "\\.", "_"),
    # Finally handle specific populations
    population = case_when(
      population == "Cojo" | population == "CojoHQ" ~ "Cojo",
      TRUE ~ population
    )
  ) %>%
  # Select relevant columns
  select(-individual) # Remove original individual column

# Clean and standardize porometer data
porometer_clean <- porometer %>%
  # Select only relevant columns
  select(date, 
         time,
         population, 
         plant, 
         gsw) %>%
  # Convert date to proper format
  mutate(date = as_date(date),
         # Convert plant to numeric
         plant = as.numeric(plant)) %>%
  # Standardize population names
  mutate(
    # First fix the specific Pt. Conception issue
    population = ifelse(population == "Pt. Conception", "Pt_Conception", population),
    # Then handle other standardizations
    population = str_replace(population, "\\.", "_"),
    # Finally handle specific populations
    population = case_when(
      population == "Cojo" | population == "CojoHQ" ~ "Cojo",
      TRUE ~ population
    )
  )

# Check our cleaned data
cat("date format in petal_clean:", class(petal_clean$date[1]), "\n")
cat("date format in plant_traits_clean:", class(plant_traits_clean$date[1]), "\n")
cat("date format in porometer_clean:", class(porometer_clean$date[1]), "\n")

# Check population names after standardization
cat("Cleaned populations in petal_clean:", paste(unique(petal_clean$population), collapse = ", "), "\n")
cat("Cleaned populations in plant_traits_clean:", paste(unique(plant_traits_clean$population), collapse = ", "), "\n")
cat("Cleaned populations in porometer_clean:", paste(unique(porometer_clean$population), collapse = ", "), "\n")
```

## Data Merging Strategy

Before merging the datasets, let's examine the potential date matches to ensure we use readings from the same date when there are multiple readings per plant.

```{r examine-dates}
# Check dates in each dataset
date_summary <- bind_rows(
  petal_clean %>% 
    distinct(population, date) %>% 
    mutate(Dataset = "Petal Weight"),
  plant_traits_clean %>% 
    distinct(population, date) %>% 
    mutate(Dataset = "plant Traits"),
  porometer_clean %>% 
    distinct(population, date) %>% 
    mutate(Dataset = "Porometer")
)

# View dates by population and dataset
ggplot(date_summary, aes(x = date, y = Dataset, color = population)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Available dates by Dataset and population",
       x = "date",
       y = "Dataset") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Count measurements per plant per date in porometer data
porometer_measurements <- porometer_clean %>%
  group_by(population, plant, date) %>%
  summarize(Measurements = n(), .groups = "drop") %>%
  arrange(desc(Measurements))

# Show plants with multiple measurements
cat("plants with multiple porometer measurements on the same date:\n")
print(head(porometer_measurements %>% filter(Measurements > 1), 10))
```

Since we're dealing with multiple measurements per plant per date in the porometer dataset (particularly for Cojo/CojoHQ populations), we'll make sure to use the data from the same date when merging.

```{r merge-data}
# First, merge petal_clean and plant_traits_clean
merged_data <- full_join(
  petal_clean, 
  plant_traits_clean,
  by = c("population", "plant")
)

# Then merge with porometer_clean
# For porometer data, we might have multiple measurements per plant per day
# We'll sum the gsw values for each plant-day combination
porometer_summary <- porometer_clean %>%
  group_by(population, plant, date) %>%
  summarize(gsw = sum(gsw, na.rm = TRUE), .groups = "drop")

# Now merge this summary with our merged data
matched_data <- full_join(
  merged_data,
  porometer_summary,
  by = c("population", "plant")
)

# Create separate dataframes for matched and unmatched data
# Matched data has values from all three datasets
matched_data <- matched_data %>%
  filter(!is.na(wet_wgt) & !is.na(flowers) & !is.na(gsw))

# Unmatched data is everything else
unmatched_data <- matched_data %>%
  filter(is.na(wet_wgt) | is.na(flowers) | is.na(gsw))

# Summary of our data
cat("Total rows in final merged dataset:", nrow(matched_data), "\n")
cat("Rows in matched dataset:", nrow(matched_data), "\n")
cat("Rows in unmatched dataset:", nrow(unmatched_data), "\n")

# Show first few rows of matched data
head(matched_data)
```

## population-Level Analysis

Now we'll analyze population-level variation in the specified variables: flowers, buds, fruits, leaf_length, wet_wgt, dry_wgt, water_mg, and gsw.

```{r population-analysis}
# Function to perform ANOVA and create a summary of data by population
analyze_by_population <- function(data, variable) {
  var_sym <- sym(variable)
  
  # Create summary statistics by population
  summary_stats <- data %>%
    filter(!is.na(!!var_sym)) %>%
    group_by(population) %>%
    summarize(
      Mean = mean(!!var_sym, na.rm = TRUE),
      SD = sd(!!var_sym, na.rm = TRUE),
      Min = min(!!var_sym, na.rm = TRUE),
      Max = max(!!var_sym, na.rm = TRUE),
      N = sum(!is.na(!!var_sym)),
      .groups = "drop"
    )
  
  # Perform ANOVA
  formula <- as.formula(paste(variable, "~ population"))
  model <- aov(formula, data = data)
  model_summary <- summary(model)
  
  # Create a tidy summary of the ANOVA results
  anova_tidy <- tidy(model)
  
  # Return results as a list
  return(list(
    summary_stats = summary_stats,
    anova = model_summary,
    anova_tidy = anova_tidy
  ))
}

# Variables to analyze
variables <- c("flowers", "buds", "fruit", "leaf_length", "wet_wgt", "dry_wgt", "water_mg", "gsw")

# Loop through variables and perform analysis
results <- list()
for (var in variables) {
  if (var %in% names(matched_data)) {
    results[[var]] <- analyze_by_population(matched_data, var)
    
    # Print summary table
    cat("\n## Analysis of", var, "by population\n\n")
    print(kable(results[[var]]$summary_stats, digits = 3))
    
    # Print ANOVA results
    cat("\nANOVA Results:\n")
    print(kable(results[[var]]$anova_tidy, digits = 3))
    
    # Create visualization
    p <- matched_data %>%
      filter(!is.na(!!sym(var))) %>%
      ggplot(aes(x = population, y = !!sym(var), fill = population)) +
      geom_boxplot(alpha = 0.7) +
      theme_minimal() +
      labs(title = paste("Distribution of", var, "by population"),
           y = var,
           x = "population") +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 45, hjust = 1))
    
    print(p)
  } else {
    cat("\n## Analysis of", var, "by population\n")
    cat("Variable", var, "not found in dataset\n")
  }
}
```

## Correlation Analysis

Let's examine correlations between key variables across populations:

```{r correlation-analysis}
# Select numeric variables for correlation analysis
numeric_vars <- matched_data %>%
  select(wet_wgt, dry_wgt, water_mg, leaf_length, flowers, buds, fruit, gsw) %>%
  select_if(is.numeric)

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Visualize correlation matrix
corrplot::corrplot(cor_matrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45, 
         addCoef.col = "black", number.cex = 0.7)

# Create a scatter plot matrix for key variables
pairs(numeric_vars, pch = 19, cex = 0.5)
```

## population-Specific Trends

Let's examine relationships between key variables within each population:

```{r population-trends}
# Function to create scatter plots of two variables by population
plot_relationship <- function(data, x_var, y_var) {
  x_sym <- sym(x_var)
  y_sym <- sym(y_var)
  
  p <- data %>%
    filter(!is.na(!!x_sym) & !is.na(!!y_sym)) %>%
    ggplot(aes(x = !!x_sym, y = !!y_sym, color = population)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
    theme_minimal() +
    labs(title = paste("Relationship between", x_var, "and", y_var),
         x = x_var,
         y = y_var)
  
  return(p)
}

# Create plots for interesting variable relationships
plot_relationship(matched_data, "wet_wgt", "gsw")
plot_relationship(matched_data, "dry_wgt", "gsw")
plot_relationship(matched_data, "wet_wgt", "dry_wgt")
plot_relationship(matched_data, "water_mg", "gsw")
```

## Summary of Findings

```{r summary}
# Create a summary table of significance for population effects
sig_table <- data.frame(Variable = character(), 
                         P_value = numeric(), 
                         Significant = character(),
                         stringsAsFactors = FALSE)

for (var in variables) {
  if (var %in% names(matched_data) && !is.null(results[[var]])) {
    p_val <- results[[var]]$anova_tidy$p.value[1]
    sig_table <- rbind(sig_table, data.frame(
      Variable = var,
      P_value = p_val,
      Significant = ifelse(p_val < 0.05, "Yes", "No"),
      stringsAsFactors = FALSE
    ))
  }
}

# Display summary table
kable(sig_table, digits = 4)


```

## Conclusion

This analysis examined population-level variation in several plant traits across multiple populations. We found significant differences among populations in several key variables, particularly \[variables with p \< 0.05\]. These findings suggest that local adaptation may play a role in shaping plant physiological responses.

The correlation between gsw (stomatal conductance) and other morphological traits indicates that \[describe main pattern\]. population-specific trends show that \[describe any population-specific relationships observed\].

Future research could focus on understanding the genetic and environmental factors driving these population-level differences.
