---
title: "Poppy_water_content"
format: html
editor: visual
---

###Daytime Petal Water Content

Petal wet weight predicts dry weight.

```{r}
#| label: wetwgt.drywgt
#| warning: false
# Import required R packages.

#library("googlesheets4")
library(reshape2)
library(dplyr)
library(broom)
library(ggplot2)
library(nlme)


# Define paths to data sets. If you don't keep your data in the same directory as the code, adapt the path names.

#setwd("/Users/gretchen/Library/CloudStorage/Box-Box/1. Lab research/CApoppy/data")

#data <- read_sheet("https://docs.google.com/spreadsheets/d/1E-IBa1wWjJXVLBjhe6bojlOcMyZRfDINiFydXIVs17s/edit#gid=0")
data <- read.csv("~/GitHub/CApoppy/data/raw/Petal_water.csv", header=TRUE, stringsAsFactors=TRUE)

data<-data[-c(1,2,3,4),]

data$Time_cat <- as.factor(data$Time_cat)
#remove outlier
data <- data[-c(76),]
data <- data[-c(79),]
#convert lists to numeric variables
# data$Wet_weight_mg<-as.numeric(unlist(data$Wet_weight_mg))
# data$Dry_weight_mg<-as.numeric(unlist(data$Dry_weight_mg))
# data$Petal_area_cm2<-as.numeric(unlist(data$Petal_area_cm2))


## Simple Linear Regression.  Does Wet_weight predict dry weight?

plot(data$Dry_weight_mg, data$Wet_weight_mg, xlab="Wet weight", ylab="Dry Weight")

model <- lm(Dry_weight_mg ~ Wet_weight_mg, data=data)
summary(model)


```

Changes in petal area across time.

```{r}
#Graphs with different estimates of water content
ggplot(data, aes(x=Dry_weight_mg, y=Petal_area_cm2, col=Time_cat)) + geom_point() +
  geom_smooth(method="lm", se=FALSE)
```

###Does petal area predict water content of petals across time?

```{r}
#|echo: false
#water_mg ~ petal area
ggplot(data, aes(x=Petal_area_cm2, y=water_mg, col=Time_cat)) + geom_point() +
  geom_smooth(method="lm", se=TRUE)
```

Water content increases with petal area at all time periods. The slope of the relationship is different at 14:00 which is just prior to flowers closing.

###Does dry weight predict total water content of petal at different times of day?

```{r}
#|echo: false

data %>% group_by(Time_cat) %>% do(tidy(lm(Petal_area_cm2 ~ water_mg, data=.)))

#water_mg ~ dryweight
ggplot(data, aes(x=Dry_weight_mg, y=water_mg, col=Time_cat)) + geom_point() +
  geom_smooth(method="lm", se=TRUE)

```

Flowers with a higher dry weight contain more water in the morning than they do later in the day. Water loss across the day is greater for larger (more dry weight) flowers as seen by the decrease in slope.

### Does the petal area predict the water content per unit area of a petal?

```{r}
data %>% group_by(Time_cat) %>% do(tidy(lm(Dry_weight_mg ~ water_mg, data=.)))

#water_cm2 ~ petal area
ggplot(data, aes(x=Petal_area_cm2, y=water_cm2, col=Time_cat)) + geom_point() +
  geom_smooth(method="lm", se=TRUE)
```

No, as petal area increases, the amount of water per unit area decreases at all times of day.

###Does dry weight of a petal predict the amount of water per unit area of that petal?

```{r}
#|echo: false
#|
data %>% group_by(Time_cat) %>% do(tidy(lm(Petal_area_cm2 ~ water_cm2, data=.)))

#water_cm2 ~ dryweight
ggplot(data, aes(x=Dry_weight_mg, y=water_cm2, col=Time_cat)) + geom_point() +
  geom_smooth(method="lm", se=TRUE)

#data %>% group_by(Time_cat) %>% do(tidy(lm(Dry_weight_mg ~ water_cm2, data=.)))

#anova
one.way <- aov(water_mg ~ Time_cat, data = data)

summary(one.way)


```

As dry weight of a petal increases, the amount of water per unit area in that petal increases at all time periods. At closing, there is a significantly higher amount of water per unit area for all sizes of petal.
