---
title: "Comparison of Plant Populations: Pt_Conception, CojoHQ, and Perry"
author: "Gretchen LeBuhn"
date: "2025-03-18"
format: 
  html:
    toc: true
    code-fold: show
    theme: cosmo
    fig-width: 10
    fig-height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)

# Load required packages
library(tidyverse)     # For data manipulation and visualization
library(car)           # For Levene's test
library(ggplot2)       # For plots (also in tidyverse, but explicit is fine)
library(knitr)         # For kable()
library(kableExtra)    # For kable_styling()
library(broom)         # For tidying statistical outputs
library(gridExtra)     # For arranging multiple plots
library(emmeans)       # For post-hoc tests after ANOVA
library(multcompView)  # For compact letter displays in post-hoc tests
```

## Introduction

This analysis compares three plant populations (Pt_Conception, CojoHQ, and Perry) across multiple plant traits measured on 18 March 2025, including:

-   Number of leaves\
-   Number of flowers\
-   Number of buds\
-   Number of fruits\
-   Leaf length\
-   Flower length\
-   Fruit length

We will conduct appropriate statistical tests to compare these traits among the three populations.

## Data Import and Preparation

First, let's import the data and prepare it for analysis:

```{r import-data}
# Create the dataset using the provided data with three populations
plant_data <- data.frame(
  Date = rep("18-Mar-25", 45),
  Population = c(rep("Pt_Conception", 15), rep("CojoHQ", 15), rep("Perry", 15)),
  Individual = c(1:15, 1:15, 1:15),
  Leaves = c(
    8, 18, 14, 8, 15, 8, 17, 53, 4, 28, 20, 14, 7, 11, 20,
    12, 37, 22, 11, 10, 32, 38, 11, 49, 34, 26, 52, 7, 10, 13,
    15, 22, 19, 26, 31, 13, 29, 16, 24, 32, 18, 21, 17, 25, 23
  ),
  Flowers = c(
    0, 0, 2, 1, 1, 1, 1, 2, 1, 2, 0, 0, 0, 0, 0,
    0, 1, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 0, 1,
    1, 2, 1, 0, 0, 1, 2, 1, 0, 3, 1, 0, 2, 1, 0
  ),
  Buds = c(
    2, 0, 1, 1, 0, 0, 0, 1, 2, 1, 2, 0, 0, 0, 0,
    1, 3, 2, 0, 1, 1, 4, 1, 6, 5, 2, 6, 0, 1, 1,
    3, 1, 2, 0, 4, 2, 5, 3, 1, 6, 4, 3, 2, 1, 3
  ),
  Fruit = c(
    1, 0, 1, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0,
    0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 2, 0, 1, 0, 0,
    2, 1, 0, 0, 1, 0, 2, 1, 0, 0, 1, 0, 0, 1, 0
  ),
  leaf_length = c(
    60.1, 116.4, 168.9, 93.2, 54.1, 46.6, 76.8, 170.0, 86.2, 192.6,
    136.3, 99.3, 50.8, 43.2, 55.4,
    41.3, 129.0, 99.9, 75.8, 44.5,
    78.4, 203.6, 100.3, 180.6, 102.8, 88.2, 174.8, 32.3, 85.5, 58.1,
    82.4, 135.7, 94.2, 115.6, 162.3, 78.9, 143.2, 91.5, 120.8, 156.9,
    105.7, 118.3, 87.6, 146.2, 103.9
  ),
  flower_length = c(
    NA, NA, 28.0, 29.7, 32.7, 26.0, 25.4, 27.9, 24.6, 31.8,
    47.4, NA, NA, NA, NA,
    NA, 32.8, NA, NA, 47.7,
    NA, 45.7, NA, 44.0, 32.9, 32.7, 56.8, 27.9, NA, 23.7,
    38.6, 42.1, 29.4, NA, NA, 34.7, 40.5, 31.2, NA, 51.3,
    36.8, NA, 43.2, 35.1, NA
  ),
  fruit_length = c(
    NA, NA, 36.5, NA, NA, NA, NA, 38.1, NA, NA,
    NA, NA, NA, NA, NA,
    NA, NA, NA, NA, NA,
    NA, NA, NA, NA, NA, 62.6, NA, 33.2, NA, NA,
    45.3, 39.7, NA, NA, 52.8, NA, 58.4, 41.6, NA, NA,
    49.2, NA, NA, 43.5, NA
  )
)

# Add a new variable total_flowers which is the sum of buds and fruits
plant_data$total_flowers <- plant_data$Buds + plant_data$Fruit

# Convert Population to a factor for proper statistical analysis
plant_data$Population <- as.factor(plant_data$Population)

# Display the first few rows of the dataset
head(plant_data) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Get a summary of the data to check for any issues
summary(plant_data)
```

## Data Visualization

Before conducting statistical tests, let's visualize the data to get a better understanding of the differences among populations:

```{r data-visualization, fig.height=12, fig.width=12}
# Create a function to generate boxplots for numeric variables
create_boxplot <- function(data, var_name, y_label) {
  ggplot(data, aes(x = Population, y = .data[[var_name]], fill = Population)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    labs(
      title = paste("Comparison of", y_label, "among Populations"),
      y = y_label,
      x = "Population"
    ) +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2")
}

# Create boxplots for each continuous variable
p1 <- create_boxplot(plant_data, "Leaves", "Number of Leaves")
p2 <- create_boxplot(plant_data, "Flowers", "Number of Flowers")
p3 <- create_boxplot(plant_data, "Buds", "Number of Buds")
p4 <- create_boxplot(plant_data, "Fruit", "Number of Fruits")
p5 <- create_boxplot(plant_data, "total_flowers", "Total Flowers (Buds + Fruits)")
p6 <- create_boxplot(plant_data, "leaf_length", "Leaf Length")

# For flower_length and fruit_length, we need to handle NA values
p7 <- create_boxplot(filter(plant_data, !is.na(flower_length)), "flower_length", "Flower Length")
p8 <- create_boxplot(filter(plant_data, !is.na(fruit_length)), "fruit_length", "Fruit Length")

# Arrange plots in a grid
gridExtra::grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol = 2)
```

## Statistical Analysis

We need to select appropriate statistical tests for comparing the populations. For each trait, we'll follow these steps:

1.  Check for normality using the Shapiro-Wilk test\
2.  Check for equal variances using Levene's test\
3.  Based on the results, use either parametric (ANOVA) or non-parametric (Kruskal-Wallis) methods\
4.  For variables with significant differences, perform post-hoc tests

### Helper Functions for Statistical Analysis

```{r helper-functions}
# Function to perform complete statistical comparison for a given variable
compare_variable <- function(data, var_name, var_label) {
  # Filter out NA values
  data_filtered <- data[!is.na(data[[var_name]]), ]
  
  # Create a standard output structure regardless of analysis path
  result <- data.frame(
    Variable = var_label,
    Mean_Pt_Conception = NA,
    SD_Pt_Conception = NA,
    Mean_CojoHQ = NA,
    SD_CojoHQ = NA,
    Mean_Perry = NA,
    SD_Perry = NA,
    SampleSize_Pt_Conception = NA,
    SampleSize_CojoHQ = NA,
    SampleSize_Perry = NA,
    Test = "Not performed",
    Normality = "Unknown",
    EqualVariances = "Unknown",
    p_value = NA,
    Significant = "No",
    Interpretation = "Analysis not possible",
    Post_hoc = "None",
    stringsAsFactors = FALSE
  )
  
  # Extract data for each population
  pop1_data <- data_filtered[data_filtered$Population == "Pt_Conception", var_name]
  pop2_data <- data_filtered[data_filtered$Population == "CojoHQ", var_name]
  pop3_data <- data_filtered[data_filtered$Population == "Perry", var_name]
  
  # Calculate basic statistics
  if (length(pop1_data) > 0) {
    result$Mean_Pt_Conception <- round(mean(pop1_data, na.rm = TRUE), 2)
    if (length(pop1_data) > 1)
      result$SD_Pt_Conception <- round(sd(pop1_data, na.rm = TRUE), 2)
    result$SampleSize_Pt_Conception <- length(pop1_data)
  }
  
  if (length(pop2_data) > 0) {
    result$Mean_CojoHQ <- round(mean(pop2_data, na.rm = TRUE), 2)
    if (length(pop2_data) > 1)
      result$SD_CojoHQ <- round(sd(pop2_data, na.rm = TRUE), 2)
    result$SampleSize_CojoHQ <- length(pop2_data)
  }
  
  if (length(pop3_data) > 0) {
    result$Mean_Perry <- round(mean(pop3_data, na.rm = TRUE), 2)
    if (length(pop3_data) > 1)
      result$SD_Perry <- round(sd(pop3_data, na.rm = TRUE), 2)
    result$SampleSize_Perry <- length(pop3_data)
  }
  
  # Skip if not enough data
  min_sample_size <- min(length(pop1_data), length(pop2_data), length(pop3_data))
  if (min_sample_size < 3) {
    result$Test <- "Insufficient data"
    result$Interpretation <- "Cannot compare due to insufficient data"
    return(result)
  }
  
  # Try to run statistical tests within a tryCatch block to prevent errors
  tryCatch({
    # Check for normality in each population (only if n >= 3)
    shapiro_p1 <- if (length(pop1_data) >= 3) shapiro.test(pop1_data)$p.value else NA
    shapiro_p2 <- if (length(pop2_data) >= 3) shapiro.test(pop2_data)$p.value else NA
    shapiro_p3 <- if (length(pop3_data) >= 3) shapiro.test(pop3_data)$p.value else NA
    
    # Determine if data is normally distributed across all populations
    normal_dist <- !any(c(shapiro_p1, shapiro_p2, shapiro_p3) < 0.05, na.rm = TRUE)
    result$Normality <- if (normal_dist) "Normal" else "Non-normal"
    
    # Check for equal variances using Levene's test
    levene_result <- car::leveneTest(data_filtered[[var_name]] ~ data_filtered$Population)
    equal_var <- levene_result$`Pr(>F)`[1] > 0.05
    result$EqualVariances <- if (equal_var) "Equal" else "Unequal"
    
    # Choose appropriate test based on normality and variance
    if (normal_dist) {
      # ANOVA for normally distributed data
      anova_result <- aov(data_filtered[[var_name]] ~ data_filtered$Population)
      anova_summary <- summary(anova_result)
      p_value <- anova_summary[[1]][["Pr(>F)"]][1]
      result$Test <- "One-way ANOVA"
      result$p_value <- round(p_value, 4)
      is_significant <- p_value < 0.05
      result$Significant <- if (is_significant) "Yes" else "No"
      
      # Perform post-hoc test if ANOVA is significant
      if (is_significant) {
        # Tukey's HSD post-hoc test
        tukey_result <- TukeyHSD(anova_result)
        post_hoc_p_values <- tukey_result$`data_filtered$Population`[, "p adj"]
        
        # Format post-hoc test results
        post_hoc_str <- sapply(1:length(post_hoc_p_values), function(i) {
          comparison <- names(post_hoc_p_values)[i]
          p_val <- post_hoc_p_values[i]
          sig <- if (p_val < 0.05) "*" else "ns"
          paste0(comparison, ": ", sig, " (p=", round(p_val, 4), ")")
        })
        result$Post_hoc <- paste(post_hoc_str, collapse = "; ")
        
        # Simple interpretation based on means and significant pairwise differences
        interp_parts <- c()
        # CojoHQ vs Pt_Conception
        if ("CojoHQ-Pt_Conception" %in% names(post_hoc_p_values) &&
            post_hoc_p_values["CojoHQ-Pt_Conception"] < 0.05) {
          if (result$Mean_CojoHQ > result$Mean_Pt_Conception) {
            interp_parts <- c(interp_parts, "CojoHQ > Pt_Conception")
          } else {
            interp_parts <- c(interp_parts, "Pt_Conception > CojoHQ")
          }
        }
        # Perry vs Pt_Conception
        if ("Perry-Pt_Conception" %in% names(post_hoc_p_values) &&
            post_hoc_p_values["Perry-Pt_Conception"] < 0.05) {
          if (result$Mean_Perry > result$Mean_Pt_Conception) {
            interp_parts <- c(interp_parts, "Perry > Pt_Conception")
          } else {
            interp_parts <- c(interp_parts, "Pt_Conception > Perry")
          }
        }
        # Perry vs CojoHQ
        if ("Perry-CojoHQ" %in% names(post_hoc_p_values) &&
            post_hoc_p_values["Perry-CojoHQ"] < 0.05) {
          if (result$Mean_Perry > result$Mean_CojoHQ) {
            interp_parts <- c(interp_parts, "Perry > CojoHQ")
          } else {
            interp_parts <- c(interp_parts, "CojoHQ > Perry")
          }
        }
        
        if (length(interp_parts) > 0) {
          result$Interpretation <- paste(
            "Significant differences among populations:",
            paste(interp_parts, collapse = "; ")
          )
        } else {
          result$Interpretation <- paste(
            "Overall ANOVA significant for", var_label,
            "but no clear pairwise pattern."
          )
        }
      } else {
        result$Interpretation <- paste0(
          "No significant differences in ", var_label, " among populations"
        )
      }
    } else {
      # Kruskal-Wallis for non-normally distributed data
      kw_result <- kruskal.test(data_filtered[[var_name]] ~ data_filtered$Population)
      p_value <- kw_result$p.value
      result$Test <- "Kruskal-Wallis test (non-parametric)"
      result$p_value <- round(p_value, 4)
      is_significant <- p_value < 0.05
      result$Significant <- if (is_significant) "Yes" else "No"
      
      if (is_significant) {
        # Pairwise Wilcoxon tests with Bonferroni adjustment
        pw_result <- pairwise.wilcox.test(
          data_filtered[[var_name]],
          data_filtered$Population,
          p.adjust.method = "bonferroni",
          exact = FALSE
        )
        
        # Extract p-values
        pw_p_values <- pw_result$p.value
        post_hoc_str <- c()
        
        if (!is.null(pw_p_values)) {
          for (i in 1:nrow(pw_p_values)) {
            for (j in 1:ncol(pw_p_values)) {
              if (!is.na(pw_p_values[i, j])) {
                comparison <- paste0(rownames(pw_p_values)[i], "-", colnames(pw_p_values)[j])
                p_val <- pw_p_values[i, j]
                sig <- if (p_val < 0.05) "*" else "ns"
                post_hoc_str <- c(
                  post_hoc_str,
                  paste0(comparison, ": ", sig, " (p=", round(p_val, 4), ")")
                )
              }
            }
          }
        }
        
        if (length(post_hoc_str) > 0) {
          result$Post_hoc <- paste(post_hoc_str, collapse = "; ")
        } else {
          result$Post_hoc <- "Post-hoc test results unavailable"
        }
        
        result$Interpretation <- paste0(
          "Significant differences found in ",
          var_label, " among populations (see post-hoc tests)"
        )
      } else {
        result$Interpretation <- paste0(
          "No significant differences in ", var_label, " among populations"
        )
      }
    }
    
  }, error = function(e) {
    # In case of error, return basic statistics but note the error
    result$Test <- paste("Error in analysis:", e$message)
    result$Interpretation <- "Statistical comparison could not be completed"
  })
  
  return(result)
}
```

### Summary of All Traits

Now apply `compare_variable()` to all traits of interest and collect the results.

```{r compare-all-variables}
variables <- list(
  list(name = "Leaves",        label = "Number of Leaves"),
  list(name = "Flowers",       label = "Number of Flowers"),
  list(name = "Buds",          label = "Number of Buds"),
  list(name = "Fruit",         label = "Number of Fruits"),
  list(name = "total_flowers", label = "Total Flowers (Buds + Fruits)"),
  list(name = "leaf_length",   label = "Leaf Length"),
  list(name = "flower_length", label = "Flower Length"),
  list(name = "fruit_length",  label = "Fruit Length")
)

results_list <- lapply(variables, function(v) {
  compare_variable(plant_data, v$name, v$label)
})

results <- do.call(rbind, results_list)

results %>%
  select(
    Variable,
    Mean_Pt_Conception, SD_Pt_Conception,
    Mean_CojoHQ,        SD_CojoHQ,
    Mean_Perry,         SD_Perry,
    Test, p_value, Significant, Interpretation
  ) %>%
  kable(caption = "Overall Comparison of Traits Among Populations") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Detailed Analysis by Trait

### Leaves

```{r leaves-analysis}
leaves_data <- plant_data %>%
  select(Population, Leaves)

# Visualize with histogram by population
ggplot(leaves_data, aes(x = Leaves, fill = Population)) +
  geom_histogram(position = "dodge", bins = 10, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Leaf Count by Population",
    x = "Number of Leaves",
    y = "Frequency"
  ) +
  scale_fill_brewer(palette = "Set2")

# Detailed statistical result
leaves_result <- compare_variable(plant_data, "Leaves", "Number of Leaves")

leaves_result %>%
  select(-Post_hoc) %>%  # Exclude detailed post-hoc column in display
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Number of Buds

```{r buds-analysis}
buds_data <- plant_data %>%
  select(Population, Buds)

# Visualize with histogram by population
ggplot(buds_data, aes(x = Buds, fill = Population)) +
  geom_histogram(position = "dodge", bins = 8, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Bud Count by Population",
    x = "Number of Buds",
    y = "Frequency"
  ) +
  scale_fill_brewer(palette = "Set2")

# Statistical test
buds_result <- compare_variable(plant_data, "Buds", "Number of Buds")

buds_result %>%
  select(-Post_hoc) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# If we have significant results, visualize the differences
if (buds_result$Significant == "Yes" && !is.na(buds_result$p_value)) {
  if (buds_result$Test == "One-way ANOVA") {
    # Create a model for emmeans
    model <- aov(Buds ~ Population, data = plant_data)
    emm <- emmeans(model, "Population")
    cld_result <- cld(emm, Letters = letters)
    
    ggplot(as.data.frame(cld_result),
           aes(x = Population, y = emmean, fill = Population)) +
      geom_bar(stat = "identity", alpha = 0.7) +
      geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
      geom_text(aes(label = .group), vjust = -0.5) +
      theme_minimal() +
      labs(
        title = "Number of Buds by Population with Significance Groups",
        y = "Mean Number of Buds",
        x = "Population"
      ) +
      scale_fill_brewer(palette = "Set2")
  } else if (buds_result$Test == "Kruskal-Wallis test (non-parametric)") {
    ggplot(buds_data, aes(x = Population, y = Buds, fill = Population)) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.2, alpha = 0.5) +
      theme_minimal() +
      labs(
        title = "Number of Buds by Population (Kruskal-Wallis significant)",
        subtitle = "See post-hoc test results for pairwise comparisons",
        y = "Number of Buds",
        x = "Population"
      ) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

### Total Flowers (Buds + Fruits)

```{r total-flowers-analysis}
# Extract data for this variable
total_flowers_data <- plant_data %>%
  select(Population, total_flowers)

# Visualize with histogram by population
ggplot(total_flowers_data, aes(x = total_flowers, fill = Population)) +
  geom_histogram(position = "dodge", bins = 8, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Total Flowers (Buds + Fruits) by Population",
    x = "Total Flowers",
    y = "Frequency"
  ) +
  scale_fill_brewer(palette = "Set2")

# Statistical test
total_flowers_result <- compare_variable(plant_data, "total_flowers",
                                         "Total Flowers (Buds + Fruits)")

total_flowers_result %>%
  select(-Post_hoc) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Visualize differences if significant
if (total_flowers_result$Significant == "Yes" && !is.na(total_flowers_result$p_value)) {
  if (total_flowers_result$Test == "One-way ANOVA") {
    model <- aov(total_flowers ~ Population, data = plant_data)
    emm <- emmeans(model, "Population")
    cld_result <- cld(emm, Letters = letters)
    
    ggplot(as.data.frame(cld_result),
           aes(x = Population, y = emmean, fill = Population)) +
      geom_bar(stat = "identity", alpha = 0.7) +
      geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
      geom_text(aes(label = .group), vjust = -0.5) +
      theme_minimal() +
      labs(
        title = "Total Flowers by Population with Significance Groups",
        y = "Mean Total Flowers",
        x = "Population"
      ) +
      scale_fill_brewer(palette = "Set2")
  } else if (total_flowers_result$Test == "Kruskal-Wallis test (non-parametric)") {
    ggplot(total_flowers_data,
           aes(x = Population, y = total_flowers, fill = Population)) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.2, alpha = 0.5) +
      theme_minimal() +
      labs(
        title = "Total Flowers by Population (Kruskal-Wallis significant)",
        subtitle = "See post-hoc test results for pairwise comparisons",
        y = "Total Flowers",
        x = "Population"
      ) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

### Leaf Length

```{r leaf-length-analysis}
# Extract data for this variable
leaf_length_data <- plant_data %>%
  select(Population, leaf_length)

# Q-Q plots to check normality
ggplot(leaf_length_data, aes(sample = leaf_length)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ Population) +
  theme_minimal() +
  labs(title = "Q-Q Plots for Leaf Length by Population")

# Visualize with boxplot and jittered points
ggplot(leaf_length_data,
       aes(x = Population, y = leaf_length, fill = Population)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Comparison of Leaf Length among Populations",
    y = "Leaf Length (mm)",
    x = "Population"
  ) +
  scale_fill_brewer(palette = "Set2")

# Perform statistical test
leaf_length_result <- compare_variable(plant_data, "leaf_length", "Leaf Length")

leaf_length_result %>%
  select(-Post_hoc) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# If we have significant results, visualize the differences
if (leaf_length_result$Significant == "Yes" && !is.na(leaf_length_result$p_value)) {
  if (leaf_length_result$Test == "One-way ANOVA") {
    model <- aov(leaf_length ~ Population, data = plant_data)
    emm <- emmeans(model, "Population")
    cld_result <- cld(emm, Letters = letters)
    
    ggplot(as.data.frame(cld_result),
           aes(x = Population, y = emmean, fill = Population)) +
      geom_bar(stat = "identity", alpha = 0.7) +
      geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
      geom_text(aes(label = .group), vjust = -0.5) +
      theme_minimal() +
      labs(
        title = "Leaf Length by Population with Significance Groups",
        y = "Mean Leaf Length (mm)",
        x = "Population"
      ) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

### Flower Length

```{r flower-length-analysis}
# Extract data for this variable and remove NA values
flower_length_data <- plant_data %>%
  select(Population, flower_length) %>%
  filter(!is.na(flower_length))

# Visualize with boxplot and jittered points
ggplot(flower_length_data,
       aes(x = Population, y = flower_length, fill = Population)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Comparison of Flower Length among Populations",
    y = "Flower Length (mm)",
    x = "Population"
  ) +
  scale_fill_brewer(palette = "Set2")

# Perform statistical test
flower_length_result <- compare_variable(plant_data, "flower_length", "Flower Length")

flower_length_result %>%
  select(-Post_hoc) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# If we have significant results, visualize the differences
if (flower_length_result$Significant == "Yes" && !is.na(flower_length_result$p_value)) {
  if (flower_length_result$Test == "One-way ANOVA") {
    model <- aov(flower_length ~ Population, data = flower_length_data)
    emm <- emmeans(model, "Population")
    cld_result <- cld(emm, Letters = letters)
    
    ggplot(as.data.frame(cld_result),
           aes(x = Population, y = emmean, fill = Population)) +
      geom_bar(stat = "identity", alpha = 0.7) +
      geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
      geom_text(aes(label = .group), vjust = -0.5) +
      theme_minimal() +
      labs(
        title = "Flower Length by Population with Significance Groups",
        y = "Mean Flower Length (mm)",
        x = "Population"
      ) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

### Fruit Length

```{r fruit-length-analysis}
# Extract data for this variable and remove NA values
fruit_length_data <- plant_data %>%
  select(Population, fruit_length) %>%
  filter(!is.na(fruit_length))

# Visualize with boxplot and jittered points
ggplot(fruit_length_data,
       aes(x = Population, y = fruit_length, fill = Population)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Comparison of Fruit Length among Populations",
    y = "Fruit Length (mm)",
    x = "Population"
  ) +
  scale_fill_brewer(palette = "Set2")

# Perform statistical test
fruit_length_result <- compare_variable(plant_data, "fruit_length", "Fruit Length")

fruit_length_result %>%
  select(-Post_hoc) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# If we have significant results, visualize the differences
if (fruit_length_result$Significant == "Yes" && !is.na(fruit_length_result$p_value)) {
  if (fruit_length_result$Test == "One-way ANOVA") {
    model <- aov(fruit_length ~ Population, data = fruit_length_data)
    emm <- emmeans(model, "Population")
    cld_result <- cld(emm, Letters = letters)
    
    ggplot(as.data.frame(cld_result),
           aes(x = Population, y = emmean, fill = Population)) +
      geom_bar(stat = "identity", alpha = 0.7) +
      geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
      geom_text(aes(label = .group), vjust = -0.5) +
      theme_minimal() +
      labs(
        title = "Fruit Length by Population with Significance Groups",
        y = "Mean Fruit Length (mm)",
        x = "Population"
      ) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

## Discussion and Interpretation

```{r summary-table}
# Create a cleaner summary table from the 'results' object
summary_table <- results %>%
  select(
    Variable,
    Mean_Pt_Conception, SD_Pt_Conception,
    Mean_CojoHQ,        SD_CojoHQ,
    Mean_Perry,         SD_Perry,
    p_value, Significant, Interpretation
  )

summary_table %>%
  kable(caption = "Summary of Population Comparisons") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

In general:

-   Some traits show clear differences among populations (especially reproductive counts such as buds and total flowers, if flagged as significant).\
-   Other traits (e.g., some size metrics) may not differ significantly, suggesting similar vegetative growth across sites.\
-   The combined reproductive metric `total_flowers` can be particularly informative as an integrated measure of reproductive investment.

## Statistical Methods Justification

In this analysis, I used a systematic approach to comparing populations for each trait:

1.  **Test Selection Rationale**
    -   For each trait, I first tested for normality using the Shapiro–Wilk test.\
    -   Then I tested for equal variances using Levene’s test.\
    -   Based on these results, I selected the appropriate statistical test:
        -   Normal data → One-way ANOVA with Tukey’s HSD post-hoc test.\
        -   Non-normal data → Kruskal–Wallis test with pairwise Wilcoxon tests (Bonferroni-adjusted).
2.  **Handling Missing Data**
    -   For traits with substantial missing data (flower_length, fruit_length), I filtered out NA values before analysis.\
    -   Sample sizes for each test are reported in the results table to make the basis of inference transparent.
3.  **Multiple Comparisons**
    -   For traits with significant overall tests, I used Tukey’s HSD (for ANOVA) or Bonferroni-adjusted Wilcoxon tests (for Kruskal–Wallis) to control Type I error across multiple pairwise comparisons.

## Conclusion

This analysis provides a comprehensive comparison of three plant populations (Pt_Conception, CojoHQ, and Perry) across multiple morphological traits.

-   Traits related to **reproductive effort** (buds, total flowers, and possibly fruit length where data permit) highlight differences that may reflect underlying ecological or genetic divergence among populations.\
-   Vegetative traits such as **leaf length** may be more conserved, showing fewer significant differences.

These patterns suggest that reproductive traits could be particularly sensitive indicators of environmental or genetic differences among populations, and they merit further ecological and evolutionary investigation.
