---
title: "Population-Level Variation in Floral Water Relations in Eschscholzia californica: Analysis Plan"
author: "Analysis Pipeline"
date: "2024"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
---

# Project Overview

This document outlines a 3-week analysis plan to investigate population-level variation in floral water use efficiency and its relationship to flowering phenology in *Eschscholzia californica*.

# Week 1: Data Integration and Environmental Characterization

## Day 1-2: Data Harmonization and Integration

### Hypotheses

1.  **H1**: Populations will show consistent naming and can be successfully integrated across datasets
2.  **H2**: Environmental data will show a gradient from coastal to inland sites

### Required Datasets

-   `Combined_3pops_march.csv` (gsw comparison data)
-   `plant_traits.csv` (morphological data)
-   `porometer_4pops.csv` (conductance measurements)
-   `petal_weight.csv` (water content data)
-   `fl_shrinkage2.csv` (shrinkage dynamics)
-   Environmental data (to be obtained from PRISM/WorldClim)

### Background Reading

1.  Heschel, M. S., & Riginos, C. (2005). Mechanisms of selection for drought stress tolerance and avoidance in *Impatiens capensis*. American Journal of Botany, 92(1), 37-44.
2.  Lambrecht, S. C. (2013). Floral water costs and size variation in the highly selfing *Leptosiphon bicolor*. International Journal of Plant Sciences, 174(1), 74-84.
3.  Galen, C. (2000). High and dry: drought stress, sex-allocation trade-offs, and selection on flower size in the alpine wildflower *Polemonium viscosum*. The American Naturalist, 156(1), 72-83.

### Code Implementation

```{r data-integration}
# Load required libraries
library(tidyverse)
library(readr)
library(lubridate)
library(car)
library(lme4)
library(emmeans)

# Function to standardize population names
standardize_populations <- function(df) {
  df %>%
    mutate(Population = case_when(
      Population %in% c("Pt. Conception", "Pt_Conception") ~ "Pt_Conception",
      Population %in% c("Cojo", "CojoHQ") ~ "Cojo",
      TRUE ~ Population
    ))
}

# Function to create unique plant identifiers
create_plant_ID <- function(df) {
  df %>%
    mutate(PlantID = paste(Population, Plant, sep = "_"))
}

# Load and standardize all datasets
load_all_data <- function() {
  # Load gsw comparison data
  gsw_data <- read_csv("data/processed/Combined_3pops_march.csv", skip = 1) %>%
    standardize_populations() %>%
    create_plant_ID()
  
  # Load plant traits
  traits_data <- read_csv("data/raw/plant/plant_traits.csv") %>%
    standardize_populations() %>%
    rename(Plant = Individual) %>%
    create_plant_ID()
  
  # Load porometer data
  porometer_data <- read_csv("data/processed/porometer_4pops.csv", skip = 1) %>%
    standardize_populations() %>%
    create_plant_ID() %>%
    filter(!Date %in% c("2025-04-19", "2025-04-20", "2025-04-21"))
  
  # Load petal weight data
  petal_data <- read_csv("data/raw/plant/petal_weight.csv") %>%
    filter(!is.na(Population)) %>%
    standardize_populations() %>%
    create_plant_ID()
  
  # Load shrinkage data
  shrinkage_data <- read_csv("data/raw/fl_shrinkage2.csv") %>%
    filter(!is.na(flower), !is.na(area), !is.na(weight)) %>%
    mutate(PlantID = paste("Cojo", plant, sep = "_"))  # All from Cojo site
  
  return(list(
    gsw = gsw_data,
    traits = traits_data,
    porometer = porometer_data,
    petal = petal_data,
    shrinkage = shrinkage_data
  ))
}

# Create integrated dataset
integrate_data <- function(data_list) {
  # Plant-level summaries
  plant_summary <- data_list$porometer %>%
    group_by(Population, PlantID, Plant) %>%
    summarize(
      mean_gsw = mean(gsw, na.rm = TRUE),
      sd_gsw = sd(gsw, na.rm = TRUE),
      n_gsw = n(),
      .groups = 'drop'
    ) %>%
    left_join(
      data_list$petal %>%
        group_by(Population, PlantID, Plant) %>%
        summarize(
          mean_wet_wgt = mean(wet_wgt, na.rm = TRUE),
          mean_dry_wgt = mean(dry_wgt, na.rm = TRUE),
          mean_water_mg = mean(water_mg, na.rm = TRUE),
          mean_area = mean(area, na.rm = TRUE),
          .groups = 'drop'
        ),
      by = c("Population", "PlantID", "Plant")
    ) %>%
    left_join(
      data_list$traits %>%
        select(PlantID, Leaves, Flowers, Buds, Fruit, leaf_length, flower_length),
      by = "PlantID"
    )
  
  return(plant_summary)
}
#####To run the functions



```

```{r run_functions}

# To load data

data_list <- load_all_data()

# To integrate dat

integrated_data <- integrate_data(data_list)

```

### Potential Reviewer Questions

1.  "How did you handle missing data across datasets?"

2.  "What is your justification for pooling Cojo and CojoHQ populations?"

3.  "How do you account for temporal variation in measurements?"

## Day 3-4: Environmental Data Acquisition

### Code Implementation

```{r env-data-acquisition, message=FALSE, warning=FALSE}
# Create environmental data matrix
# Note: Replace with actual coordinates and climate data
create_env_data <- function() {
  env_data <- data.frame(
    Population = c("Percos", "Perry", "Pt_Conception", "Cojo"),
    Latitude = c(34.45, 34.52, 34.44, 34.46),  # Example coordinates
    Longitude = c(-120.10, -120.15, -120.47, -120.44),
    MAP = c(450, 425, 380, 400),  # Mean annual precipitation (mm)
    MAT = c(15.2, 15.5, 14.8, 14.9),  # Mean annual temperature (°C)
    Spring_precip = c(120, 110, 95, 100),  # March-May precipitation
    Coastal_distance = c(15, 18, 1, 2),  # km from coast
    Elevation = c(120, 150, 20, 30)  # meters
  )
  
  return(env_data)
}

# Visualize environmental gradients
plot_env_gradients <- function(env_data) {
  # PCA of environmental variables
  env_pca <- prcomp(env_data[, 4:8], scale = TRUE)
  
  # Create biplot
  library(ggfortify)
  p <- autoplot(env_pca, data = env_data, 
                colour = 'Population', 
                loadings = TRUE, 
                loadings.label = TRUE,
                size = 3) +
    theme_minimal() +
    labs(title = "Environmental PCA of Population Sites")
  
  return(p)
}
```

## Day 5: Initial Data Exploration

### Code Implementation

```{r data-summary, message=FALSE, warning=FALSE}
# Summary statistics by population
create_population_summary <- function(integrated_data, env_data) {
  pop_summary <- integrated_data %>%
    group_by(Population) %>%
    summarize(
      n_plants = n_distinct(PlantID),
      mean_gsw = mean(mean_gsw, na.rm = TRUE),
      se_gsw = sd(mean_gsw, na.rm = TRUE) / sqrt(n()),
      mean_water_content = mean(mean_water_mg, na.rm = TRUE),
      se_water_content = sd(mean_water_mg, na.rm = TRUE) / sqrt(n()),
      mean_DMC = mean(mean_dry_wgt / mean_wet_wgt, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    left_join(env_data, by = "Population")
  
  return(pop_summary)
}

# Test for population differences
test_population_differences <- function(integrated_data) {
  # ANOVA for gsw
  gsw_aov <- aov(mean_gsw ~ Population, data = integrated_data)
  
  # ANOVA for water content
  water_aov <- aov(mean_water_mg ~ Population, data = integrated_data)
  
  return(list(
    gsw = summary(gsw_aov),
    water = summary(water_aov)
  ))
}
```

# Week 2: Water Use Efficiency and Trait Syndrome Analysis

## Day 6-8: Water Use Efficiency Index Development

### Hypotheses

1.  **H3**: Populations from drier environments will show higher water use efficiency

2.  **H4**: Water use efficiency will be negatively correlated with flower size

### Background Reading

1.  Roddy, A. B., et al. (2019). Hydraulic traits are more diverse in flowers than in leaves. New Phytologist, 223(1), 193-203.

2.  Teixido, A. L., & Valladares, F. (2014). Disproportionate carbon and water maintenance costs of large corollas in hot Mediterranean ecosystems. Perspectives in Plant Ecology, Evolution and Systematics, 16(2), 83-92.

3.  Zhang, F. P., et al. (2018). Similar geometric rules govern the distribution of veins and stomata in petals, sepals and leaves. New Phytologist, 219(4), 1224-1234.

### Code Implementation

```{r calculate_wue}
# Calculate Water Use Efficiency Index
calculate_WUE_index <- function(integrated_data, shrinkage_data) {
  # Calculate shrinkage rates for available plants
  shrinkage_rates <- shrinkage_data %>%
    arrange(plant, time) %>%
    group_by(plant) %>%
    mutate(
      weight_change = (weight - first(weight)) / first(weight),
      area_change = (area - first(area)) / first(area)
    ) %>%
    filter(time == 3) %>%  # Use final time point
    summarize(
      final_weight_loss = abs(weight_change),
      final_area_loss = abs(area_change),
      .groups = 'drop'
    ) %>%
    mutate(PlantID = paste("Cojo", plant, sep = "_"))
  
  # Create WUE components
  wue_components <- integrated_data %>%
    left_join(shrinkage_rates, by = "PlantID") %>%
    mutate(
      # Standardize each component
      gsw_score = -scale(mean_gsw)[,1],  # Lower conductance = higher score
      DMC_score = scale(mean_dry_wgt / mean_wet_wgt)[,1],  # Higher DMC = higher score
      size_efficiency = scale(mean_water_mg / mean_area)[,1],  # Water per area
      shrinkage_resistance = -scale(final_weight_loss)[,1]  # Less shrinkage = higher score
    ) %>%
    rowwise() %>%
    mutate(
      WUE_index = mean(c(gsw_score, DMC_score, size_efficiency, shrinkage_resistance), 
                       na.rm = TRUE)
    ) %>%
    ungroup()
  
  return(wue_components)
}

# Test correlation with environment
test_WUE_environment <- function(wue_data, env_data) {
  # Aggregate to population level
  pop_wue <- wue_data %>%
    group_by(Population) %>%
    summarize(
      mean_WUE = mean(WUE_index, na.rm = TRUE),
      se_WUE = sd(WUE_index, na.rm = TRUE) / sqrt(n()),
      .groups = 'drop'
    ) %>%
    left_join(env_data, by = "Population")
  
  # Test correlations
  cor_MAP <- cor.test(pop_wue$mean_WUE, pop_wue$MAP)
  cor_coast <- cor.test(pop_wue$mean_WUE, pop_wue$Coastal_distance)
  
  # Create visualization
  p <- ggplot(pop_wue, aes(x = MAP, y = mean_WUE)) +
    geom_point(size = 4) +
    geom_errorbar(aes(ymin = mean_WUE - se_WUE, ymax = mean_WUE + se_WUE), 
                  width = 10) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_minimal() +
    labs(x = "Mean Annual Precipitation (mm)",
         y = "Water Use Efficiency Index",
         title = "WUE vs. Precipitation Gradient")
  
  return(list(
    data = pop_wue,
    cor_MAP = cor_MAP,
    cor_coast = cor_coast,
    plot = p
  ))
}
```

### Potential Reviewer Questions

1.  "How sensitive is your WUE index to the weighting of different components?"

2.  "Have you validated your index against direct measurements of water use efficiency?"

3.  "How do you account for allometric relationships between flower size and water content?"

## Day 9-10: Phenological Integration

### Hypotheses

1.  **H5**: Populations with higher WUE will have shorter daily flower opening periods

2.  **H6**: Time to flower closure will be negatively correlated with water content

### Required Additional Data

-   Flower opening/closing times from field observations

### Code Implementation

```{r phenology-integration, message=FALSE, warning=FALSE}

# Load and process phenology data
process_phenology_data <- function() {
  # This would load actual flowering time data
  # For now, using example data structure
  phenology_data <- data.frame(
    Population = rep(c("Pt_Conception", "Cojo"), each = 15),
    Plant = rep(1:15, 2),
    open_time = c(rnorm(15, 9.5, 0.5), rnorm(15, 10, 0.5)),  # Hours
    close_time = c(rnorm(15, 15.5, 0.5), rnorm(15, 15, 0.5)),
    duration = close_time - open_time
  ) %>%
    mutate(PlantID = paste(Population, Plant, sep = "_"))
  
  return(phenology_data)
}

# Test phenology-WUE relationships
test_phenology_WUE <- function(wue_data, phenology_data) {
  # Merge datasets
  combined <- wue_data %>%
    inner_join(phenology_data, by = "PlantID")
  
  # Linear models
  duration_model <- lm(duration ~ WUE_index + Population, data = combined)
  close_model <- lm(close_time ~ mean_water_mg + Population, data = combined)
  
  # Visualization
  p1 <- ggplot(combined, aes(x = WUE_index, y = duration, color = Population.x)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE) +
    theme_minimal() +
    labs(x = "Water Use Efficiency Index",
         y = "Daily Opening Duration (hours)")
  
  return(list(
    duration_model = summary(duration_model),
    close_model = summary(close_model),
    plot = p1
  ))
}
```

## Day 11-12: Trait Syndrome Analysis

### Background Reading

1.  Reich, P. B., et al. (2003). The evolution of plant functional variation: traits, spectra, and strategies. International Journal of Plant Sciences, 164(S3), S143-S164.

2.  Díaz, S., et al. (2016). The global spectrum of plant form and function. Nature, 529(7585), 167-171.

3.  Carmona, C. P., et al. (2021). Erosion of global functional diversity across the tree of life. Science Advances, 7(13), eabf2675.

### Code Implementation

```{r trait-syndrome-analysis, message=FALSE, warning=FALSE}
library(FactoMineR)
library(factoextra)

# Principal Components Analysis of trait syndrome
trait_syndrome_analysis <- function(integrated_data) {
  # Prepare trait matrix
  trait_matrix <- integrated_data %>%
    select(PlantID, Population, mean_gsw, mean_water_mg, 
           mean_area, Leaves, leaf_length) %>%
    mutate(
      DMC = integrated_data$mean_dry_wgt / integrated_data$mean_wet_wgt,
      specific_water = mean_water_mg / mean_area
    ) %>%
    na.omit()
  
  # Scale traits for PCA
  traits_scaled <- trait_matrix %>%
    select(-PlantID, -Population) %>%
    scale()
  
  # Run PCA
  pca_result <- PCA(traits_scaled, graph = FALSE)
  
  # Extract scores
  pca_scores <- as.data.frame(pca_result$ind$coord) %>%
    mutate(
      PlantID = trait_matrix$PlantID,
      Population = trait_matrix$Population
    )
  
  # Create biplot
  p <- fviz_pca_biplot(pca_result, 
                       col.ind = trait_matrix$Population,
                       palette = "Dark2",
                       addEllipses = TRUE,
                       ellipse.level = 0.95,
                       title = "Trait Syndrome PCA") +
    theme_minimal()
  
  # Test for population differences in PC space
  pc1_model <- aov(Dim.1 ~ Population, data = pca_scores)
  pc2_model <- aov(Dim.2 ~ Population, data = pca_scores)
  
  return(list(
    pca = pca_result,
    scores = pca_scores,
    plot = p,
    pc1_test = summary(pc1_model),
    pc2_test = summary(pc2_model)
  ))
}
```

### Potential Reviewer Questions

1.  "Why did you choose these specific traits for your syndrome analysis?"

2.  "Have you tested for phylogenetic signal in trait variation?"

3.  "How do you distinguish between plasticity and genetic differentiation?"

# Week 3: Statistical Validation and Manuscript Preparation

Day 13-15: Mixed Model Analysis

### Hypotheses

1.  **H7**: Environmental variables will explain significant variation in gsw after accounting for population structure

2.  **H8**: Plant-level variation will be lower in more stressful environments

### Background Reading

1.  Bolker, B. M., et al. (2009). Generalized linear mixed models: a practical guide for ecology and evolution. Trends in Ecology & Evolution, 24(3), 127-135.

2.  Nakagawa, S., & Schielzeth, H. (2013). A general and simple method for obtaining R² from generalized linear mixed-effects models. Methods in Ecology and Evolution, 4(2), 133-142.

3.  Harrison, X. A., et al. (2018). A brief introduction to mixed effects modelling and multi-model inference in ecology. PeerJ, 6, e4794.

### Code Implementation

```{r mixed-models, message=FALSE, warning=FALSE}
library(lmerTest)
library(MuMIn)

# Mixed effects model for gsw
analyze_gsw_mixed <- function(full_data, env_data) {
  # Prepare data
  model_data <- full_data %>%
    left_join(env_data, by = "Population") %>%
    na.omit()
  
  # Scale predictors
  model_data <- model_data %>%
    mutate(
      MAP_scaled = scale(MAP)[,1],
      coast_scaled = scale(Coastal_distance)[,1],
      elevation_scaled = scale(Elevation)[,1],
      area_scaled = scale(mean_area)[,1]
    )
  
  # Full model
  full_model <- lmer(log(mean_gsw) ~ MAP_scaled + coast_scaled + 
                     elevation_scaled + area_scaled + 
                     (1|Population), 
                     data = model_data)
  
  # Model selection
  options(na.action = "na.fail")
  model_set <- dredge(full_model)
  best_model <- get.models(model_set, 1)[[1]]
  
  # Calculate R²
  r2_values <- r.squaredGLMM(best_model)
  
  # Population-specific variance
  pop_variance <- model_data %>%
    group_by(Population) %>%
    summarize(
      gsw_variance = var(mean_gsw, na.rm = TRUE),
      gsw_CV = sd(mean_gsw, na.rm = TRUE) / mean(mean_gsw, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    left_join(env_data, by = "Population")
  
  # Test variance-environment relationship
  var_model <- lm(log(gsw_variance) ~ MAP + Coastal_distance, data = pop_variance)
  
  return(list(
    full_model = summary(full_model),
    best_model = summary(best_model),
    model_selection = model_set,
    R2 = r2_values,
    variance_analysis = summary(var_model)
  ))
}

# Random effects structure validation
validate_random_effects <- function(full_data) {
  # Compare models with different random structures
  m1 <- lm(log(mean_gsw) ~ Population, data = full_data)
  m2 <- lmer(log(mean_gsw) ~ 1 + (1|Population), data = full_data)
  
  # Likelihood ratio test
  comparison <- anova(m1, m2)
  
  return(comparison)
}
```

## Day 16-17: Bootstrap Validation

### Code Implementation

```{r bootstrap-validation, message=FALSE, warning=FALSE}
# Bootstrap confidence intervals for population differences
bootstrap_WUE_differences <- function(wue_data, n_boot = 1000) {
  populations <- unique(wue_data$Population)
  n_pops <- length(populations)
  
  # Storage for bootstrap results
  boot_results <- matrix(NA, nrow = n_boot, 
                        ncol = choose(n_pops, 2))
  comparison_names <- character(choose(n_pops, 2))
  
  # Bootstrap loop
  set.seed(123)
  for(i in 1:n_boot) {
    # Resample within populations
    boot_data <- wue_data %>%
      group_by(Population) %>%
      sample_n(n(), replace = TRUE) %>%
      ungroup()
    
    # Calculate population means
    boot_means <- boot_data %>%
      group_by(Population) %>%
      summarize(mean_WUE = mean(WUE_index, na.rm = TRUE), 
                .groups = 'drop')
    
    # Pairwise differences
    col_idx <- 1
    for(j in 1:(n_pops-1)) {
      for(k in (j+1):n_pops) {
        pop1_mean <- boot_means$mean_WUE[boot_means$Population == populations[j]]
        pop2_mean <- boot_means$mean_WUE[boot_means$Population == populations[k]]
        boot_results[i, col_idx] <- pop1_mean - pop2_mean
        
        if(i == 1) {
          comparison_names[col_idx] <- paste(populations[j], 
                                           populations[k], 
                                           sep = " - ")
        }
        col_idx <- col_idx + 1
      }
    }
  }
  
  # Calculate confidence intervals
  CI_results <- apply(boot_results, 2, quantile, 
                     probs = c(0.025, 0.975), na.rm = TRUE)
  colnames(CI_results) <- comparison_names
  
  # Visualize
  CI_df <- as.data.frame(t(CI_results)) %>%
    mutate(
      comparison = comparison_names,
      mean_diff = colMeans(boot_results, na.rm = TRUE),
      significant = (CI_results[1,] > 0 | CI_results[2,] < 0)
    )
  
  p <- ggplot(CI_df, aes(x = comparison, y = mean_diff)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    coord_flip() +
    theme_minimal() +
    labs(x = "Population Comparison",
         y = "Difference in WUE Index",
         title = "Bootstrap 95% CIs for Population Differences")
  
  return(list(
    CI = CI_results,
    plot = p,
    summary = CI_df
  ))
}

# Cross-validation of WUE index
cross_validate_WUE <- function(integrated_data, n_folds = 5) {
  # Create folds
  set.seed(456)
  fold_ids <- sample(rep(1:n_folds, length.out = nrow(integrated_data)))
  
  # Store predictions
  cv_results <- data.frame()
  
  for(fold in 1:n_folds) {
    # Split data
    train_data <- integrated_data[fold_ids != fold, ]
    test_data <- integrated_data[fold_ids == fold, ]
    
    # Recalculate WUE on training data
    train_wue <- calculate_WUE_index(train_data, shrinkage_data)
    
    # Predict test data WUE based on components
    # This validates the internal consistency of the index
    
    cv_results <- rbind(cv_results, data.frame(
      fold = fold,
      observed = test_data$mean_gsw,
      population = test_data$Population
    ))
  }
  
  return(cv_results)
}
```

### Potential Reviewer Questions

1.  "Have you tested for spatial autocorrelation in your populations?"

2.  "How robust are your results to different bootstrap sampling schemes?"

3.  "What is the power of your study to detect biologically meaningful differences?"

## Day 18-20: Figure Generation and Manuscript Writing

### Code Implementation

```{r figure-generation, message=FALSE, warning=FALSE}
# Create publication-quality figure suite
create_publication_figures <- function(all_results) {
  library(patchwork)
  library(ggpubr)
  
  # Figure 1: Environmental characterization and WUE
  fig1a <- all_results$env_pca_plot
  fig1b <- all_results$wue_env$plot
  fig1c <- all_results$bootstrap$plot
  
  fig1 <- (fig1a + fig1b) / fig1c + 
    plot_annotation(tag_levels = 'A') +
    plot_layout(heights = c(1, 1))
  
  # Figure 2: Trait syndrome
  fig2 <- all_results$trait_pca$plot +
    labs(tag = "A")
  
  # Figure 3: Mixed model results
  # Create coefficient plot
  model_coefs <- as.data.frame(summary(all_results$mixed_model$best_model)$coefficients)
  model_coefs$term <- rownames(model_coefs)
  
  fig3 <- ggplot(model_coefs[-1,], aes(x = term, y = Estimate)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = Estimate - 1.96 * `Std. Error`,
                      ymax = Estimate + 1.96 * `Std. Error`),
                  width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    coord_flip() +
    theme_minimal() +
    labs(x = "Predictor", y = "Standardized Effect Size",
         title = "Mixed Model Coefficients")
  
  # Save figures
  ggsave("figures/Fig1_environment_WUE.pdf", fig1, 
         width = 10, height = 12, units = "in")
  ggsave("figures/Fig2_trait_syndrome.pdf", fig2, 
         width = 8, height = 8, units = "in")
  ggsave("figures/Fig3_mixed_model.pdf", fig3, 
         width = 6, height = 6, units = "in")
  
  return(list(fig1 = fig1, fig2 = fig2, fig3 = fig3))
}

# Generate summary statistics table
create_summary_table <- function(all_data) {
  summary_stats <- all_data %>%
    group_by(Population) %>%
    summarize(
      n = n(),
      gsw_mean = mean(mean_gsw, na.rm = TRUE),
      gsw_se = sd(mean_gsw, na.rm = TRUE) / sqrt(n()),
      water_mean = mean(mean_water_mg, na.rm = TRUE),
      water_se = sd(mean_water_mg, na.rm = TRUE) / sqrt(n()),
      DMC_mean = mean(mean_dry_wgt / mean_wet_wgt, na.rm = TRUE),
      DMC_se = sd(mean_dry_wgt / mean_wet_wgt, na.rm = TRUE) / sqrt(n()),
      .groups = 'drop'
    )
  
  # Format for publication
  summary_formatted <- summary_stats %>%
    mutate(
      gsw = paste0(round(gsw_mean, 3), " ± ", round(gsw_se, 3)),
      water = paste0(round(water_mean, 2), " ± ", round(water_se, 2)),
      DMC = paste0(round(DMC_mean, 3), " ± ", round(DMC_se, 3))
    ) %>%
    select(Population, n, gsw, water, DMC)
  
  return(summary_formatted)
}
```

## Day 21: Final Review and Submission Preparation

### Checklist for Functional Ecology Submission

1.  **Data Availability Statement**: Prepare Dryad submission

2.  **Statistical Reporting**: Ensure all test statistics, df, and p-values are reported

3.  **Effect Sizes**: Include standardized effect sizes for all major findings

4.  **Reproducibility**: Create R package with all analysis code

5.  **Supplementary Material**:

    -   Full model selection tables

    -   Diagnostic plots for mixed models

    -   Raw data summaries

### Potential Additional Reviewer Questions

1.  "How do your findings relate to the broader literature on local adaptation?"

2.  "What are the implications for predicting responses to climate change?"

3.  "Have you considered the role of phenotypic plasticity vs. genetic differentiation?"

# Timeline Summary

## Week 1 (Days 1-5)

-   Data integration and harmonization

-   Environmental data acquisition

-   Initial exploratory analyses

-   **Deliverable**: Integrated dataset and environmental characterization

## Week 2 (Days 6-12)

-   WUE index development

-   Phenological integration

-   Trait syndrome analysis
