---
title: "Cal Poppy ovary temp"
author: "Gretchen LeBuhn"
date: "5/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

These are data collected to track flowering time in Eschscholzia californica.  This represents two nights on the first night, we held on flower open using wire.

```{r set libraries and dir, echo=FALSE}
library(lubridate) # for working with dates
library(ggplot2)  # for creating graphs
library(scales)   # to access breaks/formatting functions
library(gridExtra) # for arranging plots
library(tidyr) # for reshaping data
library(dplyr)


```

```{r, set dir, echo = FALSE}
setwd('~/Dropbox/CApoppy/data')
#datadir="~/Dropbox/CApoppy/data"
```
I've modified the CR6 data by dropping the last header line and switching lines 2 and 3 so that line 2 becomes the header for the data.  I had to do text to columns and then save it as .csv for it to import correctly.
```{r, import data, echo = FALSE}

air_temp = read.csv("~/Dropbox/CApoppy/data/air_temp/55000000492FED21_052422.csv", skip= 14, header = T )
ov_temp = read.csv("~/Dropbox/CApoppy/data/fl_temp/cr6_Ten_min.csv", header = T, stringsAsFactors=FALSE)


#format date and time
air_temp$DateTime <- as.POSIXct(paste(air_temp$Date, air_temp$Time, sep=" "), format="%Y-%m-%d %I:%M:%S %p", 
                     tz = "UTC")
ov_temp$DateTime <- as.POSIXct(ov_temp$TIMESTAMP, format="%m/%d/%y %I:%M", tz = "UTC")





df1 <- merge(air_temp, ov_temp, by = 'DateTime', all=TRUE)
df1 %>% select(-Date, -Time, -Unit,-Time2, -TIMESTAMP,-RECORD)

df2 <- df1 %>%
  select(DateTime, Value, PTemp_C, Temp_C.1., Temp_C.2. ) %>%
  gather(key = "variable", value = "value", -DateTime)
head(df2, 3)
df2 <-na.omit(df2)




```

## Plotting temperature data

Using this plotting tutorial
https://www.neonscience.org/resources/learning-hub/tutorials/dc-time-series-plot-ggplot-r

```{r airtemp, echo=FALSE}
#plot(air_temp$Time,air_temp$Value)
#ggplot(aes(x = DateTime, y = Value, colour=Date), data = air_temp) + geom_line()

AirTemp10min <- ggplot(air_temp, aes(DateTime, Value)) +
           geom_point(na.rm=TRUE, color="blue", size=3, pch=18) +
          ggtitle("Temp every 10 min May 22-24, 2-22") +
           xlab("Date") + ylab("Air Temperature (C)")

#render plot
AirTemp10min

AirTemp10minb <- AirTemp10min + 
  (scale_x_datetime(breaks=date_breaks("6 hours"),labels=date_format("%I %M")))

#render plot
AirTemp10minb




AllTemp <- ggplot(df2, aes(DateTime, value)) +
           geom_line(aes(colour=variable, group=variable)) + # colour, group both depend on variable
           geom_point(aes(colour=variable),                # colour depends on variable
               size=1)+
         ggtitle("Temp every 10 min May 22-24, 2-22") +
         xlab("Date") + ylab("Air Temperature (C)")

#render plot
AllTemp
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
